\documentclass[11pt]{article}

\usepackage[preprint]{acl}

\usepackage{times}
\usepackage{latexsym}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{inconsolata}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{cuted}
\usepackage{capt-of}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,backgrounds,calc,patterns}

\newtheorem{proposition}{Proposition}

\title{AlignAtt4LLM: Fast AlignAtt for Decoder-Only LLMs \\ in an IWSLT 2026 Simultaneous Speech Translation System}
\author{Anonymous ACL Submission}

\usepackage{cleveref} % for \Cref{tab:abc} instead of guessing what it is and then Section~\ref{tab:abc}

\def\Aref#1{Appendix~\ref{#1}}
\def\AAName{AlignAtt4LLM}
\newcommand{\offlinecell}[1]{\textcolor{gray}{\emph{#1}}}

\begin{document}
\maketitle

\begin{abstract}
We describe \AAName{}, an IWSLT 2026 simultaneous speech translation system
for English-to-German, English-to-Italian, and English-to-Chinese. The system
is a synchronous cascade: Qwen3-ASR with forced alignment produces an
incrementally updated source transcript with word end times, and Gemma-4 E4B-it
translates that prefix under an MT-side AlignAtt policy.

To our knowledge, this is the first application of AlignAtt to a decoder-only
LLM, where the encoder-decoder cross-attention used by earlier AlignAtt
systems is absent. We recover a usable policy by proposing (1) an explicit source
span in the prompt, (2) offline selection of translation-specific
alignment heads, (3) selective \emph{qk-fast} replay of the draft-to-source
attention block, and (4) runtime query/key capture that preserves
model outputs bit-identically.

On the IWSLT 2026 development set, \AAName{} reaches a low-latency operating
point near two seconds of compute-unaware LongYAAL and trades 1.32 to 1.53
additional seconds for clear gains in BLEU and XCOMET-XL in the high-latency
regime. Head selection remains an end-to-end quality/latency trade-off, and
the deployed \emph{qk-fast} path keeps the MT observer cheap enough to run in
the same serialized GPU as ASR.
\end{abstract}

\section{Introduction}
\label{sec:intro}

This paper describes the \AAName{} IWSLT 2026 submission system for
English-to-German, English-to-Italian, and English-to-Chinese simultaneous
speech translation \citep{anastasopoulos-etal-2026-findings}. The system is a
synchronous cascade: Qwen3-ASR with the
Qwen3 forced aligner \citep{qwen3asr26} produces an incrementally updated
source transcript with word end times, and Gemma-4 E4B-it \citep{gemma4card26}
translates the current source prefix under an MT-side AlignAtt policy. We keep
two latency-quality regimes, one tuned near the low-latency boundary and one
tuned for higher translation quality.

The central contribution is to make an offline decoder-only LLM usable as an
online MT module. Standard AlignAtt reads encoder-decoder cross-attention, but
Gemma-4 has no such cross-attention. We instead expose the source transcript as
a known prompt span, select a small set of translation-specific self-attention
heads offline, and accept only draft prefixes whose reconstructed attention
signal remains within the currently available source frontier.

The second contribution is making the implementation fast enough for
computation-aware mode. Both ASR and MT are served through vLLM \citep{vllm23},
which gives the cascade one high-throughput inference stack and hot model reuse
across chunks. On the MT side, this also means the policy cannot inspect a
Python-visible attention matrix: self-attention is hidden inside fused kernels.
We therefore capture the exact query/key tensors
consumed by the deployed Gemma attention module and replay only the
draft-to-source block needed by AlignAtt.

We will release a public version of the implementation with the camera-ready
version.

\Cref{sec:background} gives the task and prior-work context,
\Cref{sec:system} describes the submitted cascade,
\Cref{sec:method} details the MT-side AlignAtt realization, and
\Cref{sec:results} reports the evaluation.

\section{Background}
\label{sec:background}

\textbf{AlignAtt} \citep{alignatt23,streamatt24} is a simultaneous policy that
lets an offline model operate in streaming mode. It was recently applied in the
state-of-the-art simultaneous system \citep{cuniiwslt25} at IWSLT 2025. At each generation step, the
policy reads an attention-derived source position for the next target token and
stops generation when that position is not yet safely available. Previous
AlignAtt systems obtained this signal from encoder-decoder cross-attention.

\textbf{Offline models with online policies.} Repurposing offline models for
simultaneous translation is attractive because it reuses high-quality,
multilingual, instruction-following models without training a new streaming
architecture for every latency regime. The cost is that an offline model is not
trained to decide when a partial source prefix is sufficient, and naive
re-translation can flicker or hallucinate on incomplete input. Online policies
such as AlignAtt provide the missing commit decision while preserving the
offline model's quality and flexibility \citep{alignatt23,cuniiwslt25}.

\textbf{Decoder-only LLMs} have recently become central in high-quality
text-to-text MT systems \citep{kocmi-etal-2024-findings,kocmi-etal-2025-findings}.
Prior simultaneous systems have already used instruction-tuned decoder-only
LLMs by prompting them with the translation direction, current source prefix,
and already emitted target prefix \citep{cuniiwslt25}. Long context,
instruction following, and in-context examples make this family attractive for
robust MT, but previous decoder-only simultaneous implementations used
LocalAgreement \citep{polak-etal-2022-cuni,polak23_interspeech} rather than
AlignAtt. LocalAgreement is a strong simultaneous policy, but it does not
exploit model-internal attention.

\textbf{Translation-specific attention heads.} \citet{tokenheads26} show that
multilingual decoder-only LLMs contain a sparse set of heads whose attention
argmaxes track source-target word alignments. This suggests a decoder-only
AlignAtt path, but only if two runtime conditions are met: the source span must
be identifiable in the prompt, and the policy-visible attention rows
must be reconstructed from the same tensors used by the deployed inference
engine. The method below supplies both conditions.

\subsection{IWSLT 2026 Simultaneous Task}

\textbf{Simulation environment.} The IWSLT 2026 Simultaneous Task proposes the
Simulstream toolkit \citep{simulstream25} for simulating simultaneous
translation sessions. We follow the long-form, unsegmented task setup, which
allows re-translation at the evaluation interface, but \AAName{} itself emits
append-only incremental output.

\textbf{Development data.} We use the official IWSLT 2026 MCIF development
set.
We report translation quality with BLEU
\citep{papineni-etal-2002-bleu}, chrF \citep{popovic-2015-chrf}, and
XCOMET-XL \citep{xcometxl24}. Translation latency is reported with LongYAAL
\citep{longyaal25} in both compute-unaware and compute-aware variants.


\section{\AAName{} System Overview}
\label{sec:system}

\paragraph{ASR.}
The speech component of the submitted cascade uses two upstream Qwen components:
Qwen3-ASR-1.7B transcribes the live audio tail, and Qwen3-ForcedAligner-0.6B
assigns word-level start and end times to that transcript \citep{qwen3asr26}.
At every chunk boundary, the cascade re-transcribes the current live utterance
tail and invokes the forced aligner with timestamp output enabled; the word
times are therefore produced online inside our simulation run, not added later
as an offline post-processing step. Adjacent ASR hypotheses are stabilized by a
longest-common-prefix commit rule up to sentence-final punctuation, while the
remaining live tail is allowed to change on the next chunk.
We retain this dedicated aligner because it performed better than all the alternative candidates summarized in \Aref{app:appendix-asr}.
% from all candidate than another candidatethe Gemma
%AlignAtt ASR variants tested during development, and because it is the only
%tested front end that jointly provides low WER, online word-end times, and a
%comfortable real-time margin. Appendix~\ref{sec:appendix-asr} summarizes the
%ASR front-end comparison and the tail-reliability analysis that motivated a
%small future source hold-back.

\paragraph{MT.}
The translation component uses Gemma-4 E4B-it \citep{gemma4card26} served through
vLLM \citep{vllm23}.
With every source update, Gemma receives the current transcript prefix, the already accepted
target prefix, and a fixed translation instruction, then greedily generates a
draft continuation capped at 16 new tokens in the partial streaming mode used
by the submitted systems. AlignAtt accepts the longest draft prefix whose
alignment signal stays on the currently accessible side of the source frontier.

\paragraph{Streaming loop.}
The cascade is synchronous and chunk-based. Each chunk first updates the ASR
transcript prefix, and then launches one MT request. Source words are considered
accessible only once their end time is older than a hold-back margin; in the
submitted systems, this margin is 0 ms, so a word becomes accessible as
soon as its aligned end time is observed. Both submitted regimes also delay the
first MT emission until 2 seconds of source audio have accumulated.

\paragraph{Latency-quality parameters.}
We keep two official submission regimes. The low-latency submission uses
$\Delta_{\mathrm{chunk}}=850$ ms and the high-latency submission uses
$\Delta_{\mathrm{chunk}}=1500$ ms. Both use the same ASR backend, Gemma-4 MT
backbone, and MT-side AlignAtt realization. The submitted systems use
the same per-direction top-8 MT head sets, a width-7 source-axis median filter
applied before taking the source-side argmax, and a one-source-token border
margin $b=1$.



The submitted cascade itself is
synchronous and single-GPU: each chunk triggers ASR first and MT second. This
conservative regime cleanly separates policy behavior from scheduler overlap
and defines the computation-aware setting used in our submitted measurements.
\Cref{fig:system} summarizes this chunk-synchronous runtime path.

\input{generated/submitted_cascade.tex}

\Cref{fig:alignatt-substrates} summarizes the conceptual shift. The
policy intuition remains the same as in encoder-decoder AlignAtt, but the
alignment signal must now be recovered from a prompt-structured self-attention
substrate rather than read directly from cross-attention.

\input{generated/alignatt_substrates.tex}

\section{AlignAtt for Decoder-Only LLMs}
\label{sec:method}

This section focuses on the method that enables AlignAtt for decoder-only LLMs.

\subsection{Prompt Layout}

Standard AlignAtt does not directly apply because a decoder-only LLM has no
decoder cross-attention. We therefore make the source span explicit in the
causal prompt. At chunk $k$, the serialized chat prompt is a concatenation of
the system prompt, live transcript prefix, translation instruction, accepted
translation prefix, and current draft:
\begin{equation}
\mathbf{p}^{(k)} =
\left[\mathbf{p}^{\mathrm{sys}}
\Vert \mathbf{s}^{(k)}
\Vert \mathbf{p}^{\mathrm{instr}}
\Vert \mathbf{y}_{1:m_k}
\Vert \widetilde{\mathbf{y}}^{(k)}\right],
\label{eq:mt-prompt}
\end{equation}
where $\mathbf{s}^{(k)}$ is the live transcript prefix returned by ASR,
$\mathbf{y}_{1:m_k}$ is the already accepted translation prefix, and
$\widetilde{\mathbf{y}}^{(k)}$ is the draft proposed by Gemma-4 at the current
step. The transcript prefix is therefore a contiguous prompt span with a known
map $\phi^{(k)}$ from source-word indices to prompt positions.

\subsection{Selection of Alignment Heads}

Not all self-attention heads carry a useful translation signal.
Following \citet{tokenheads26}, we operationalize this
with a two-stage offline calibration procedure under the exact prompt layout of
Eq.~\eqref{eq:mt-prompt}: GPT-5-mini \citep{openai2025gpt5developers} first
provides word-level source-target alignments on held-out parallel text, then
every Gemma head is scored with Translation Score (TS), the aligned-token
argmax accuracy used by \citet{tokenheads26}, on those aligned examples. We
retain the top $k=8$ heads per language pair. This head set is
fixed at inference time and is the only part of the policy that depends on
offline calibration.
Appendix~\ref{sec:appendix-alignatt} reports the calibration details, corpus
sizes, and per-direction score summaries. The retained MT heads are sparse and
concentrated in a limited late-depth region of the backbone rather than being
spread uniformly; Appendix~\ref{sec:appendix-alignatt} further visualizes that pattern.
%in Figure~\ref{fig:mt-alignatt-heads}.

The policy only reads the source slice of each draft attention row,
$\mathbf{A}^{(\ell,h)}_{t,\phi^{(k)}(s)}$, but unlike encoder-decoder
cross-attention, this slice is not source-normalized. Off-source mass on the
accepted target prefix, prompt template, and speculative suffix is therefore
structural rather than implementation noise. On the 78 text-only qualitative
probes used for diagnostics, drafted units allocate on average 9\%
to accessible source tokens, 8\% to inaccessible source tokens, 81\% to
non-source prompt positions, and 2\% to the speculative suffix; among units
that the policy actually accepts, the inaccessible-source share drops from 8\%
to 1\%.

\subsection{Selective Reconstruction for the Policy}

The policy never needs the full $n\times n$ self-attention matrix. It only
needs draft rows against source columns. For selected heads $(\ell,h)$, draft
positions $t$, and source positions $s$, we reconstruct
\begin{equation}
\widehat{A}^{(\ell,h)}_{t,s}
=
\frac{\exp\!\left(
\gamma^{(\ell)} q^{(\ell,h)}_t \cdot k^{(\ell,h)}_{\phi^{(k)}(s)}
  + m_{t,\phi^{(k)}(s)}
\right)}
{\sum_j \exp\!\left(
\gamma^{(\ell)} q^{(\ell,h)}_t \cdot k^{(\ell,h)}_j + m_{t,j}
\right)},
\label{eq:mt-qk-fast}
\end{equation}
where $q$ and $k$ are the captured post-normalization tensors after rotary
position embedding (RoPE; \citealp{su2021roformer}) actually consumed by the
deployed attention module, $\gamma^{(\ell)}$ is that
module's runtime scaling factor, and $m_{t,j}$ replays the same causal and
sliding-window masks as the fused forward. We call this replay path
\emph{qk-fast}: it reconstructs only the policy-visible block and matches the
deployed attention algebra up to floating-point reassociation relative to the
fused forward.

\input{generated/qk_fast_selective_reconstruction.tex}

From these per-head rows we build two source-side provenance features for each
drafted token: accessible-source mass $\pi_t^{\mathrm{acc}}$ and
inaccessible-source mass $\pi_t^{\mathrm{inacc}}$,
\begin{align}
  \pi_t^{\mathrm{acc}}
  &=
  \sum_{s \in \mathcal{S}^{(k)}_{\mathrm{acc}}} \bar{\mathbf{r}}_t(s),
  \label{eq:mt-prov-acc}\\
  \pi_t^{\mathrm{inacc}}
  &=
  \sum_{s \in \mathcal{S}^{(k)}_{\mathrm{inacc}}} \bar{\mathbf{r}}_t(s),
  \label{eq:mt-provenance}
\end{align}
where $\bar{\mathbf{r}}_t$ is the head-averaged reconstructed row. We
additionally normalize per-head rows online with prefix Welford statistics and
apply a short median filter along the source axis before taking the source-side
argmax. The submitted gate only thresholds $\pi_t^{\mathrm{acc}}$; the
complementary $\pi_t^{\mathrm{inacc}}$ is retained as a diagnostic split for
Figure~\ref{fig:mt-selective-reconstruction} and the appendix-level provenance
analyses. This stabilizes the alignment peak without changing the underlying
replayed rows.

\subsection{AlignAtt Acceptance Policy}

The submitted MT policy is a first-failure scan over drafted tokens. Let
$\hat{s}_t$ be the source-side argmax of the filtered row for draft token $t$,
and let $N^{(k)}_{\mathrm{acc}}$ be the number of source words whose ASR end
time is already on the accessible side of the frontier. The scan stops when one
of three conditions first fires:
\begin{align}
  \text{\textsc{source-frontier}} :\quad
  &\hat{s}_t \ge N^{(k)}_{\mathrm{acc}} + b,
  \label{eq:mt-stop-frontier}\\
  \text{\textsc{argmax-mass-weak}} :\quad
  &\bar{\mathbf{r}}_t(\hat{s}_t) < \tau_{\mathrm{argmax}},
  \label{eq:mt-stop-argmax}\\
  \text{\textsc{provenance-weak}} :\quad
  &\pi_t^{\mathrm{acc}} < \tau_{\mathrm{src}}.
  \label{eq:mt-stop-prov}
\end{align}
In the submitted 850/1500 ms operating points, $b=1$,
$\tau_{\mathrm{argmax}}=0$, and $\tau_{\mathrm{src}}=0$, so the optional
argmax-mass and minimum-source-mass gates are present in the runtime but
inactive in the submitted operating points.

\input{generated/appendix_policy_pipeline.tex}

Figure~\ref{fig:policy-pipeline} makes the decision path explicit. One branch
averages the retained heads and sums accessible-source mass to obtain
$\pi_t^{\mathrm{acc}}$. The other normalizes and smooths the same replayed rows
before taking the source-side argmax $\hat{s}_t$. The gate accepts token $t$
if the three tests in
Eqs.~\eqref{eq:mt-stop-frontier} to~\eqref{eq:mt-stop-prov} all pass; a
left-to-right first-failure scan then emits the longest accepting draft prefix
and rounds back to the last completed stability unit. Here a stability unit is
the smallest target fragment that the tokenizer treats as safe to commit: a
whitespace-delimited lexical word in spacing scripts such as EN$\to$DE/IT, or
a single CJK character in EN$\to$ZH, so partial subword fragments are never
emitted.

\subsection{Runtime Query/Key Capture}

Equation~\eqref{eq:mt-qk-fast} is only useful if the deployed runtime exposes
the exact queries and keys consumed by attention. In the deployed vLLM path,
those tensors never appear as ordinary Python-visible objects. We therefore
install a per-layer observer, copy prompt keys and draft-row queries into
fixed-shape buffers, and route capture through a custom-op-style call that
survives runtime graph lowering. A zero-valued sentinel
dependency keeps the observer in the transitive fan-in of the graph output:
\begin{equation}
  \mathbf{o}'_\ell =
  \mathbf{o}_\ell +
  \Phi\bigl(\ell,\mathrm{pos},\mathbf{Q}^{(\ell)},\mathbf{K}^{(\ell)}\bigr),
  \qquad \Phi \equiv \mathbf{0}.
  \label{eq:mt-sentinel}
\end{equation}

\input{generated/appendix_observer_lifecycle.tex}

Figure~\ref{fig:observer-lifecycle} separates the capture path into three
stages. During setup, we patch the attention forward and install fixed-shape
observer slots before the graph is captured. During each forward, the live
observer records prompt keys and draft-row queries for the selected heads after
the module's $q/k$ normalization and rotary embedding, then returns
$\Phi \equiv \mathbf{0}$ back into the attention output. After the forward, a
short Python-side replay reconstructs only the selective
draft-to-source block consumed by the policy and by the parity suite. The
sentinel add-back of Eq.~\eqref{eq:mt-sentinel} is therefore only a liveness
device: without the additive zero, graph lowering dead-code-eliminates the
observer path and no usable Q/K buffers survive on the deployed runtime.
Appendix~\ref{sec:appendix-observer} keeps the replay equations, parity
measurements, and qualitative probe.

\section{Results}
\label{sec:results}

\subsection{Experimental Setup}
\label{sec:setup}

The submitted streaming experiments run on a single NVIDIA A40. The MT component uses Gemma-4 E4B-it
in bfloat16 through vLLM; the speech component uses Qwen3-ASR with
forced alignment. In evaluation, we use the official IWSLT 2026 MCIF dev set
($\sim 2.1$ hours total, 21 long academic talks) resegmented by OmniSTEval.
We report BLEU and chrF against the dev references, XCOMET-XL for
translation adequacy \citep{xcometxl24}, and compute-unaware and aware
LongYAAL for latency \citep{longyaal25}. The streaming loop follows the
Simulstream setting of unsegmented long-form audio with revision-capable
evaluation \citep{simulstream25}, but our submitted cascade itself is
synchronous: ASR and MT are serialized on one GPU so that policy effects are
not confounded with scheduler overlap.

For the auxiliary runtime study in Section~\ref{sec:runtime}, we compare three
MT seams on a fixed suite of 16 text-only prompts: a minimal Transformers eager
reference, a Transformers SDPA \emph{qk-fast} replay reference, and the
deployed vLLM \emph{qk-fast} path. Each seam uses greedy decoding with the
same retained head set.

\subsection{Submitted System Results}
\label{sec:main-results}

Table~\ref{tab:e2e} reports the submitted system candidates together with the
organizers' no-context baseline points supplied for comparison. The pattern is
consistent across EN$\to$DE, EN$\to$ZH, and EN$\to$IT: the low-latency regime
stays near a 2\,s CU-LongYAAL point, while the high-latency regime moves to a
clearly stronger quality point at the expected latency cost. CA-LongYAAL is
consistently below CU-LongYAAL here because OmniSTEval's computation-aware mode
replaces each chunk-boundary audio increment by the actual wall-clock increment
spent processing that chunk. Since the deployed system runs faster than real
time, the computation-aware timestamps can be earlier than the
compute-unaware chunk-boundary timestamps. The columns are therefore not
inverted. We read Table~\ref{tab:e2e} as evidence that the \AAName{}
realization supports
both a stable low-latency operating point and a higher-quality regime on the
same decoder-only Gemma backend.

\begin{table}[t]
\centering
\footnotesize
\setlength{\tabcolsep}{2.5pt}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{l l l c c c r r}
\toprule
Pair & System & Regime & BLEU & chrF & XCOMET-XL & \multicolumn{2}{c}{LongYAAL (s)} \\
\cmidrule(lr){7-8}
& & & & & & CU & CA \\
\midrule
\multirow{5}{*}{en$\to$de}
& Org. baseline & low  & $22.35$ & $56.7$ & $0.748$ & \textbf{1.81} & n/a \\
& Ours          & low  & \textbf{28.76} & \textbf{62.1} & \textbf{0.875} & $2.00$ & $1.63$ \\
& Org. baseline & high & $26.31$ & $59.2$ & $0.819$ & \textbf{2.63} & n/a \\
& Ours          & high & \textbf{32.63} & \textbf{64.2} & \textbf{0.902} & $3.53$ & $3.14$ \\
& \offlinecell{Our cascade (offline)} & \offlinecell{ref.} & \offlinecell{38.57} & \offlinecell{67.1} & \offlinecell{0.938} & \offlinecell{n/a} & \offlinecell{n/a} \\
\midrule
\multirow{5}{*}{en$\to$zh}
& Org. baseline & low  & \textbf{40.85} & $34.1$ & \textbf{0.750} & \textbf{1.91} & n/a \\
& Ours          & low  & $36.01$ & \textbf{35.0} & $0.743$ & $1.95$ & $1.77$ \\
& Org. baseline & high & \textbf{43.85} & \textbf{37.8} & \textbf{0.795} & $3.48$ & n/a \\
& Ours          & high & $39.86$ & \textbf{37.8} & $0.778$ & \textbf{3.27} & $3.09$ \\
& \offlinecell{Our cascade (offline)} & \offlinecell{ref.} & \offlinecell{48.53} & \offlinecell{43.4} & \offlinecell{0.848} & \offlinecell{n/a} & \offlinecell{n/a} \\
\midrule
\multirow{5}{*}{en$\to$it}
& Org. baseline & low  & $30.63$ & $62.0$ & $0.683$ & \textbf{1.76} & n/a \\
& Ours          & low  & \textbf{40.10} & \textbf{68.0} & \textbf{0.805} & $1.98$ & $1.62$ \\
& Org. baseline & high & $37.28$ & $65.4$ & $0.781$ & \textbf{3.30} & n/a \\
& Ours          & high & \textbf{44.46} & \textbf{70.1} & \textbf{0.841} & $3.48$ & $3.10$ \\
& \offlinecell{Our cascade (offline)} & \offlinecell{ref.} & \offlinecell{49.88} & \offlinecell{73.0} & \offlinecell{0.895} & \offlinecell{n/a} & \offlinecell{n/a} \\
\bottomrule
\end{tabular}
}
\caption{\textbf{Submitted cascade on the IWSLT 2026 dev set}
(21 clips, OmniSTEval long-form resegmentation). Ours low rows use the
submitted setting with $\Delta_{\mathrm{chunk}}=850$ ms; Ours high rows
use $\Delta_{\mathrm{chunk}}=1500$ ms. Organizer baseline rows are
the no-context baseline outputs provided with CU LongYAAL; their
computation-aware LongYAAL was not available. Offline rows are diagnostic
quality-only cascades with Qwen full-audio ASR followed by Gemma final-mode MT;
they have no streaming latency score.}
\label{tab:e2e}
\end{table}

Against the supplied organizers' no-context baselines, \AAName{} clearly wins
the $<2$\,s and $<4$\,s latency regimes for EN$\to$DE and EN$\to$IT.
EN$\to$ZH is mixed: our system reaches comparable chrF at high latency and
slightly higher chrF at low latency, but the organizers' baseline remains ahead
on BLEU and XCOMET-XL.
The offline diagnostic rows show substantial backbone headroom once online
commitment is removed. They are not latency-comparable to the submitted
streaming systems, but they help separate model capacity from the cost of the
simultaneous policy. \Aref{sec:appendix-offline} gives the minimal setup
details.

\subsection{Additional Analysis: ASR Tail Reliability}
\label{sec:asr-tail-analysis}

The submitted systems expose each Qwen3 ASR word to MT as soon as its aligned
end time is observed. A post-hoc reliability analysis suggests a slightly more
conservative default for future runs: clip the last 250\,ms of the live ASR
tail before passing the source prefix to MT. This is a policy-level timing
margin rather than a lexical repair. Because the Qwen path re-transcribes the
live utterance tail from scratch at each chunk, the final ASR words are exactly
the part most likely to be revised on the next chunk. This instability is partly
masked by AlignAtt, which often stops MT before translating the newest source
words near the frontier. Reducing tail noise could therefore allow a less
conservative acceptance policy, not only a longer hold-back. On the 21 MCIF dev talks,
the reference error rate of words at the current Qwen3 ASR tail end is 17.1\%,
but drops to 8.3\% at 250\,ms and then remains nearly flat. We therefore keep
the submitted numbers unchanged, but recommend a 250\,ms ASR-tail hold-back as
the maintained default going forward. \Aref{app:appendix-asr} gives
the comparison with the alternative ASR component parameters.

\input{generated/asr_tail_risk_curve.tex}

\subsection{Head Filtering Matters End-to-End}
\label{sec:head-filtering}

Table~\ref{tab:mt-e2e-head-filtering} is an auxiliary EN$\to$DE rerun included
only to isolate the MT head-set effect under the maintained \emph{qk-fast}
runtime. End-to-end quality is nearly equivalent, but the latency split is
informative. CU-LongYAAL increases by 100.3\,ms, which likely means the all-head
observer is slightly less precise and therefore a bit more generous on tokens
whose support extends beyond the currently available source prefix. CA-LongYAAL
rises more strongly, by +179.5\,ms, because replaying all 336 heads forces the
runtime to reconstruct much more attention state at each MT step. We therefore
read the auxiliary comparison as evidence that a small retained head set
captures most of the useful MT alignment signal at much lower runtime cost.
Appendix~\ref{sec:appendix-alignatt} gives the held-out calibration numbers
behind the retained head sets.

\input{generated/mt_e2e_head_ablation.tex}

\subsection{Attention versus Static Target Cutoff}
\label{sec:cutoff-microscope}

To test whether the attention signal is useful beyond a fixed target-side
delay, we run a controlled quality--latency diagnostic on three EN$\to$DE
development clips held out from the microscope used to inspect the policy. All
systems use the same Qwen ASR, Gemma MT model, 850\,ms chunks, and evaluator.
We change only the partial MT acceptance rule: an attention policy, or a static
policy that accepts each Gemma draft after removing the last $x$ target
stability units. Unlike the replay-only microscope, this is scored from the
actual streaming hypotheses, so the comparison lives directly in the BLEU/chrF
versus LongYAAL plane.

\input{generated/mt_cutoff_policy_heldout3_quality_latency.tex}

The held-out diagnostic is deliberately small, but it is more informative than
the single-clip frontier. A first tuned policy that required every emitted token
to carry at least 0.05 accessible-source mass looked Pareto-optimal on the
microscope clip, but failed here: most stops were caused by the
\texttt{provenance\_weak} gate, which is too strong for target function words
and morphology. We therefore use the attention signal primarily as a
future-dependence test: block tokens when the selected heads show evidence that
they still depend on inaccessible source, while keeping only a weak
accessible-source sanity check. The reported \textsc{AlignAtt} point uses top-4
heads, border margin 1, a soft frontier that ignores frontier crossings with
less than 0.03 future-source mass, an upper bound of 0.15 future-source mass,
and a small 0.005 accessible-source-mass floor.

\input{generated/mt_cutoff_policy_heldout3_frontier_gap.tex}

Table~\ref{tab:mt-cutoff-quality-latency} shows the target operating point:
\textsc{AlignAtt} stays below the 2\,s CU budget at 1.92\,s and beats every
fixed cutoff that also satisfies that budget. The best admissible fixed cutoff
is \texttt{cut-last-2}, at 26.00 BLEU, 59.94 chrF, and 1.68\,s CU; attention
raises this to 28.57 BLEU and 60.52 chrF while remaining under budget. The next
fixed rule, \texttt{cut-last-3}, does improve quality, but only after crossing
the 2\,s CU target. Thus the attention policy is not dominated by any fixed
cutoff from \texttt{cut-last-0} to \texttt{cut-last-6}; it gives a different
latency allocation rather than just choosing another static delay.

Table~\ref{tab:mt-cutoff-frontier-gap} gives a stricter interpolation check.
At matched latency, attention sits above the static frontier on BLEU
(+0.37 at CU, +0.13 at CA) and slightly below it on chrF (-0.14 at CU, -0.22
at CA). We therefore do not claim that attention uniformly dominates a
continuous interpolation of fixed cutoffs. The stronger and better-supported
claim is Pareto: a paired bootstrap over the 117 resegmented segments gives a
99.5\% weakest-case probability that no fixed cutoff dominates the attention
point in BLEU, chrF, CU, and CA. This is the mechanism we want from attention:
not a single fixed delay, but a source-conditioned cutoff that spends latency
only when the current target token appears to depend on future source context.
XCOMET-XL is left blank for this diagnostic because the local checkpoint load
was killed before scoring; the corpus-level benchmark rows above report the
metric where it is available.

\subsection{Runtime Cost of the MT Component}
\label{sec:runtime}

The MT component must require low computational resources to justify deployment.
Figure~\ref{fig:mt-capture-speed} shows that the vLLM replay path is
much cheaper than the minimal eager seam we use as an inspection reference:
median cost drops from 63.7 to 25.4 ms/token, while the Transformers
SDPA \emph{qk-fast} replay remains close to the eager baseline at
59.0 ms/token.
This confirms that the MT contribution is not merely inspectable in principle;
it is practical inside the runtime we actually deploy.

\input{generated/mt_capture_speed_benchmark.tex}

\subsection{Coordination Regimes and Compute-Aware Latency}

Our cascade loads the ASR and MT models as two separate blocking
\texttt{vllm.LLM} instances on a single GPU, which serializes every audio chunk
into an ASR decode followed by an MT decode
(Fig.~\ref{fig:coord-regimes}\,(c)). This conservative regime keeps policy
effects separate from scheduler overlap. Alternative pipelined or fully
asynchronous regimes would mainly reduce compute-aware LongYAAL at fixed
compute-unaware latency, but would require a non-blocking scheduler and a more
careful treatment of stale MT requests. We leave that scheduler axis to future
work.

\section{Conclusion}
\label{sec:conclusion}

The retained IWSLT 2026 system is a Qwen3 forced ASR + Gemma-4 MT cascade, and
its key technical ingredient is an AlignAtt policy adapted to decoder-only MT.
Deterministic prompt layout, offline head selection, selective \emph{qk-fast}
replay, and runtime query/key capture make the MT policy usable on the deployed
vLLM path. On the
dev set, this yields a low-latency operating point near 2\,s CU-LongYAAL and a
high-latency regime that is roughly 1.5\,s slower but clearly stronger in
BLEU and XCOMET-XL. Compared with the supplied organizers' no-context
baselines, the system clearly wins the $<2$\,s and $<4$\,s latency regimes for
EN$\to$DE and EN$\to$IT, while EN$\to$ZH remains mixed: chrF is competitive, but
BLEU and XCOMET-XL still favor the baseline. This appears partly tied to the Gemma-4 MT backbone,
whose Chinese generations were weaker in our runs; because \AAName{} only
requires a deterministic prompt layout, calibrated heads, and Q/K capture, the
same policy can be ported to other decoder-only LLMs, making a Qwen-family MT
backbone a natural next target for EN$\to$ZH.

\bibliography{references}

\appendix

\section{Additional ASR Analyses}
\label{app:appendix-asr}

The analyses below justify the source front end used by the submitted cascade
and the recommended source-tail default for future runs.

\paragraph{ASR front-end selection.}
We tested three ASR front ends during development: Qwen3-ASR with the Qwen3
forced aligner, Voxtral Realtime 4B, and a direct Gemma E4B ASR local-agreement
probe. Table~\ref{tab:asr-frontends} summarizes the 21-talk MCIF comparison.
Voxtral gives the lowest final WER, but it is not real-time in our serialized
single-GPU loop and its CU-LongYAAL is higher than Qwen3. It also advances the
transcript with a more regular lag, whereas Qwen3 often exposes compact
multi-word updates, which better matches our chunk-synchronous MT policy.
Gemma E4B would be attractive as a single-model substrate, but its WER, prompt
leakage on long talks, and unstable raw latency make it unsuitable here. We
therefore retain Qwen3 forced ASR.

\input{generated/asr_frontend_selection.tex}

\paragraph{ASR tail clipping.}
Even with Qwen3, words at the live tail are less reliable than words slightly
behind it. Figure~\ref{fig:asr-tail-risk} shows that most of the reference
error reduction happens within the first 250\,ms, supporting a future default
that hides the final 250\,ms from MT.

\subsection{Offline Cascade Diagnostic}
\label{sec:appendix-offline}
As a development tool that brackets the streaming numbers from above, we also
ran an offline cascade on the same 21 MCIF dev talks: Qwen3-ASR was applied
once to each complete audio file, then Gemma-4 was run in final mode on
sentence-level chunks resegmented as in the streaming evaluation. The full-audio ASR transcripts have 7.40\% weighted
corpus WER against the English reference (90\% audio-bootstrap interval
$[6.27, 8.59]$), so the offline run sees source much closer to the reference
than any online chunk. Differences between offline and streaming output
therefore isolate the cost of the AlignAtt commit policy and the live ASR tail
from the intrinsic quality of the backbone: drops on EN$\to$ZH that already
appear here point to Gemma-4 rather than to the simultaneous policy.

\section{Coordination Regimes and Additional MT Alignment-Head Diagnostics}
\label{sec:appendix-alignatt}

\input{generated/coordination_regimes.tex}

\input{generated/decoder_only_alignatt_heads.tex}

\input{generated/mt_head_filtering_ablation.tex}

\section{Observer Replay and Qualitative Diagnostics}
\label{sec:appendix-observer}

Below we report the prompt/suffix replay equations, one qualitative probe, and
the numerical parity measurements on the deployed vLLM path.

\paragraph{Prompt-space replay details.}
After capture, we replay prompt-side and suffix-side logits separately from the
stored buffers. Writing $\mathbf{B}^{(\ell,h)}_\star :=
\mathbf{B}^{(\ell)}_\star[h]$ for head-specific captured tensors,
$\gamma^{(\ell)}$ for the attention module's runtime scaling factor, and
$\mathbf{M}^{(\ell,h)}_{\mathrm{prompt}}$ /
$\mathbf{M}^{(\ell,h)}_{\mathrm{suffix}}$ for the replayed prompt-side
sliding-window and suffix-side causal/window masks, we compute
\begin{align}
  \mathbf{P}^{(\ell,h)} &=
  \gamma^{(\ell)}
  \mathbf{B}^{(\ell,h)}_{\mathrm{dQ}}
  \bigl(\mathbf{B}^{(\ell,h)}_{\mathrm{pK}}\bigr)^{\!\top}
  + \mathbf{M}^{(\ell,h)}_{\mathrm{prompt}},
  \label{eq:prompt-logits}\\
  \mathbf{S}^{(\ell,h)} &=
  \gamma^{(\ell)}
  \mathbf{B}^{(\ell,h)}_{\mathrm{dQ}}
  \bigl(\mathbf{B}^{(\ell,h)}_{\mathrm{dK}}\bigr)^{\!\top}
  + \mathbf{M}^{(\ell,h)}_{\mathrm{suffix}},
  \label{eq:suffix-logits}\\
  \widetilde{\mathbf{A}}^{(\ell,h)}
  &= \mathrm{softmax}[\mathbf{P}^{(\ell,h)}, \mathbf{S}^{(\ell,h)}].
  \label{eq:vllm-softmax}
\end{align}
Restricting $\widetilde{\mathbf{A}}^{(\ell,h)}$ to the source columns
$\phi^{(k)}(s)$ recovers the policy-visible block of
Eq.~\eqref{eq:mt-qk-fast}. The replay cost therefore scales with the selected
head set and the short draft length, not with the full $LHn^2$ attention
tensor.

\paragraph{Qualitative reconstruction example.}
Figure~\ref{fig:mt-selective-reconstruction} shows the full reconstruction
stack on an EN$\to$DE text-only MT probe: the exact prompt partition seen by
the decoder, word-level aggregation of the replayed rows for readability, and
the four-way provenance accounting for every drafted word.

\input{generated/mt_selective_reconstruction_example.tex}

The green band in Figure~\ref{fig:mt-selective-reconstruction} is previously
committed target text reused as causal context; it is not itself the object of
the gate. The orange words are the current draft scanned left-to-right by the
policy. We allow $\hat{s}_t$ to move backward within a live draft when the
decoder revisits earlier source material under reordering, which is why the
residual prompt and suffix masses remain explicit in the decoder-only
formulation.

\paragraph{Numerical parity.}
Let $\mathbf{A}^{(\mathrm{TF})}$ denote the reference attention tensor produced
by a Transformers reference forward on the identical prompt. On a curated
parity set, we observe
\begin{equation}
  \begin{aligned}
    \|\Delta\mathbf{A}\|_\infty &\leq 1.2 \times 10^{-2}, \\
    \|\Delta\mathbf{A}\|_1 / (nL_k) &\leq 4 \times 10^{-4},
  \end{aligned}
  \label{eq:parity}
\end{equation}
with bit-identical acceptance decisions between the reference and deployed paths.
The observer is therefore not just fast enough to keep, but faithful enough to
support controlled MT ablations.

\end{document}
