\documentclass[11pt,twocolumn]{article}
\usepackage[a4paper,margin=2.0cm,columnsep=0.7cm]{geometry}
\usepackage{amsmath,amssymb,bm,amsthm}
\newtheorem{proposition}{Proposition}
\usepackage[hidelinks]{hyperref}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{positioning,fit,backgrounds,arrows.meta,calc,patterns}

\title{AlignAtt without Cross-Attention:\\
  Engine-Native Attention Probing for Simultaneous LLM Translation}
\author{}
\date{}

\begin{document}
\maketitle

\begin{abstract}
We study AlignAtt for simultaneous translation with decoder-only LLMs, where
the model exposes no decoder cross-attention and modern inference stacks fuse
attention inside graph-compiled kernels. We recover source-conditioned
provenance from causal self-attention by combining a deterministic prompt
layout, offline selection of alignment heads, and exact selective
reconstruction of the draft-to-source attention block via
\emph{qk-fast}. We then describe an engine-native observer that captures the
required query and key tensors under graph-compiled inference without breaking
compilation or altering model outputs. The result is a single AlignAtt
substrate that applies to both ASR and MT roles while preserving numerical
parity with eager reference execution.
\end{abstract}

% ============================================================
% SECTION 1 — POSITIONING
% ============================================================
\section{Positioning}
\label{sec:positioning}

Attention-based policies have already proved useful for simultaneous speech
translation. AlignAtt derives audio-to-translation alignments from
encoder--decoder cross-attention at inference time \cite{alignatt23}, while
StreamAtt addresses long-form direct streaming speech translation through
attention-based audio-history selection \cite{streamatt24}. Recent
interpretability work on multilingual LLMs further identifies sparse
translation-specific alignment heads in decoder-only models
\cite{tokenheads26}. Our setting differs from all three: we study a cascade
whose MT component is a decoder-only LLM with no cross-attention, and whose
deployed runtime executes fused, graph-compiled attention.

We therefore ask whether AlignAtt can be reformulated so that it remains exact,
selective, and observable under these constraints. Our answer has two coupled
parts: a virtual source-restricted view of self-attention, and an engine-native
capture mechanism that exposes the query/key tensors required for selective
reconstruction without perturbing model outputs. For evaluation we use
Simulstream, a framework for long-form revision-capable streaming speech
translation \cite{simulstream25}, and we report compute-unaware LongYAAL on
unsegmented audio \cite{longyaal25}. All experiments use a synchronous cascade
loop in order to isolate policy behaviour from scheduler effects. The ASR and
MT components are architecturally separable and could in principle be pipelined
in future system work, but such overlap is not evaluated here.

% ============================================================
% SECTION 2 — THE CASCADE
% ============================================================
\section{Simultaneous Cascade Architecture}
\label{sec:cascade}

\paragraph{Streaming input and alignment backend.}
The system is invoked on a monotone sequence of timestamps
$t_1 < t_2 < \cdots$ with $t_{k+1} - t_k = \Delta_{\mathrm{chunk}}$
(by default $\Delta_{\mathrm{chunk}} = 850$\,ms).
At each $t_k$, an alignment backend returns a partial transcript
$\mathbf{s}^{(k)} = (s^{(k)}_1, \ldots, s^{(k)}_{N_k})$ with per-word end-times
$\theta^{(k)}_i \in \mathbb{R}_{\geq 0}$, $i = 1, \ldots, N_k$. The analysis
below only assumes this contract: an incremental recognition/alignment module
that returns a live source prefix together with per-word end-times.
Throughout, bold symbols such as $\mathbf{s}^{(k)}$ denote ordered token
sequences, while calligraphic symbols denote index sets; in particular
$\mathcal{S}^{(k)} := \{1,\ldots,N_k\}$ is the local index set of the live
source prefix at step $k$.

\paragraph{Source accessibility frontier.}
Given an inaccessibility margin $\tau_{\mathrm{in}} \geq 0$, the
\emph{accessible source index set} at time $t_k$ is
\begin{equation}
  \mathcal{S}^{(k)}_{\mathrm{acc}}
  = \Bigl\{\, i \in \mathcal{S}^{(k)}
     \;\Bigm|\; \theta^{(k)}_i \leq t_k - \tau_{\mathrm{in}} \,\Bigr\},
  \label{eq:frontier}
\end{equation}
with $N^{(k)}_{\mathrm{acc}} = |\mathcal{S}^{(k)}_{\mathrm{acc}}|$ and the
convention that on a final chunk ($t_k = T$) we set
$N^{(k)}_{\mathrm{acc}} = N_k$.
The inaccessible complement is
$\mathcal{S}^{(k)}_{\mathrm{inacc}} = \mathcal{S}^{(k)} \setminus
\mathcal{S}^{(k)}_{\mathrm{acc}}$.
These positions are conditioned on but \emph{not} allowed to drive accepted
target tokens (Sec.~\ref{sec:qk-fast}).

\paragraph{Translation backend.}
The MT backend maps the live source $\mathbf{s}^{(k)}$ and the accepted target
prefix $\mathbf{y}_{1:m_k}$ to a draft $\widetilde{\mathbf{y}}^{(k)} =
(\widetilde{y}^{(k)}_1, \ldots, \widetilde{y}^{(k)}_{L_k})$ of at most
$L_{\mathrm{draft}}$ new target tokens.

\paragraph{Acceptance and emission.}
An AlignAtt policy $\Pi$ turns the draft
$\widetilde{\mathbf{y}}^{(k)}$ into an accepted suffix
$\mathbf{a}^{(k)}$ of length $\ell_k \in \{0, 1, \ldots, L_k\}$,
\begin{equation}
  \mathbf{a}^{(k)} = \Pi\!\left(\widetilde{\mathbf{y}}^{(k)},\, \mathcal{S}^{(k)}_{\mathrm{acc}}\right),
  \label{eq:accept-prefix}
\end{equation}
so $\mathbf{a}^{(k)} = (\widetilde{y}^{(k)}_1, \ldots, \widetilde{y}^{(k)}_{\ell_k})$ is the
retained draft prefix,
and the accepted target prefix is extended monotonically,
\begin{equation}
  \mathbf{y}_{1:m_{k+1}} = \mathbf{y}_{1:m_k} \,\Vert\, \mathbf{a}^{(k)},
  \label{eq:monotone}
\end{equation}
with $m_{k+1} = m_k + \ell_k$. The invariant $m_k$ is non-decreasing;
already-emitted target tokens are never retracted.

Figure~\ref{fig:dual-alignatt-cascade} gives the operational view used in this
work: a standalone decoder-only Gemma cascade where each chunk triggers an ASR
request and an MT request, the confirmed ASR transcript prefix defines the
visible MT source span, and the accepted target prefix is reused as causal
target prefill.
\input{generated/standalone_dual_alignatt_cascade.tex}

\paragraph{Latency functional.}
Let $\tau^{\mathrm{emit}}_j$ denote the wall-clock emission time of target
token $y_j$ and $\tau^{\mathrm{src}}_j$ the end-time of the source word
responsible for it. The primary latency metric we track is the
compute-unaware LongYAAL latency \cite{longyaal25}:
\begin{equation}
  \mathrm{LongYAAL}_{\mathrm{CU}}(\mathbf{y})
  = \frac{1}{|\mathbf{y}|} \sum_{j=1}^{|\mathbf{y}|}
  \bigl(\tau^{\mathrm{emit}}_j - \tau^{\mathrm{src}}_j\bigr)_+,
  \label{eq:lya-cu}
\end{equation}
with $(u)_+ = \max(u,0)$.

% ============================================================
% SECTION 3 — ALIGNATT FOR DECODER-ONLY LLMs
% ============================================================
\section{AlignAtt for Decoder-Only LLMs}
\label{sec:alignatt-llm}

\paragraph{From encoder--decoder AlignAtt to decoder-only LLMs.}
In its original formulation, AlignAtt exploits decoder cross-attention in an
encoder--decoder speech translation model \cite{alignatt23}: each decoder row
$\mathbf{A}^{\mathrm{xattn}}_{t,\cdot}$ is normalized over encoder source
positions only, so source provenance is built into the architecture. In a
decoder-only LLM, there is no cross-attention: a single token stream of
length $n$ mixes instructions, source, conversation history, and the draft
being generated, and every row of causal self-attention
$\mathbf{A}^{\mathrm{self}}_{t,\cdot}$ is normalized over all $n$ prompt
positions. Two things are thus lost relative to the encoder--decoder setting:
\emph{(i)} the attention row no longer distinguishes source from non-source
by construction, and
\emph{(ii)} no head is architecturally reserved for source alignment.

\paragraph{Prompt-induced source span.}
We recover a source--target distinction by \emph{prompt layout} rather than
architecture. The MT prompt is assembled deterministically from a fixed system
preamble $\mathbf{p}^{\mathrm{sys}}$, the live source $\mathbf{s}^{(k)}$, a
task instruction $\mathbf{p}^{\mathrm{instr}}$ followed by the accepted target
prefix $\mathbf{y}_{1:m_k}$, and the draft $\widetilde{\mathbf{y}}^{(k)}$:
\begin{equation}
  \mathbf{p}^{(k)} = \bigl[\,
  \underbrace{\mathbf{p}^{\mathrm{sys}}}_{\mathcal{P}_1}
  \;\Vert\; \underbrace{\mathbf{s}^{(k)}}_{\mathcal{S}^{(k)}}
  \;\Vert\; \underbrace{\mathbf{p}^{\mathrm{instr}}, \mathbf{y}_{1:m_k}}_{\mathcal{P}_2}
  \;\Vert\; \underbrace{\widetilde{\mathbf{y}}^{(k)}}_{\mathcal{D}} \,\bigr],
  \label{eq:layout}
\end{equation}
where $\mathcal{S}^{(k)} = \mathcal{S}^{(k)}_{\mathrm{acc}} \sqcup \mathcal{S}^{(k)}_{\mathrm{inacc}}$
is the live source index set at decode step $k$, $\mathcal{P} = \mathcal{P}_1
\cup \mathcal{P}_2$ collects all non-source prompt positions, and
$\mathcal{D}$ is the draft. The sets $\mathcal{S}^{(k)}_{\mathrm{acc}}$ and
$\mathcal{S}^{(k)}_{\mathrm{inacc}}$ are exactly the accessibility partition
of Eq.~\eqref{eq:frontier}. Let
$\phi^{(k)} : \mathcal{S}^{(k)} \hookrightarrow \{1,\ldots,n\}$ map local
source indices to absolute prompt positions. Eq.~\eqref{eq:layout} makes the
source span explicit in prompt space even though the model still executes
ordinary causal self-attention.

\paragraph{Source-restricted attention row.}
Given this layout, the object relevant to alignment is simply the slice of
self-attention over source columns: for draft position $t \in \mathcal{D}$ and
source index $s \in \mathcal{S}^{(k)}$, we read
$\mathbf{A}^{(\ell,h)}_{t,\, \phi^{(k)}(s)}$, layer $\ell$, head $h$. Unlike
in the encoder--decoder setting, $\sum_{s \in \mathcal{S}^{(k)}}
\mathbf{A}^{(\ell,h)}_{t,\, \phi^{(k)}(s)} < 1$ in general, because the row is
normalized over the full prompt. The residual mass on non-source prompt
positions and on the draft is not noise to be discarded; in
Sec.~\ref{sec:qk-fast} it is kept available during replay, while the
source-side decomposition drives the acceptance policy.

\paragraph{What changes and what stays the same.}
Relative to standard AlignAtt, the online decision rule stays the same in
spirit: a first-failure scan accepts the longest draft prefix whose
source-side alignment signal remains on the accessible side of the source
frontier. What changes is the \emph{substrate} on which that signal is
observed. In the encoder--decoder case the alignment row is source-normalized
by construction, whereas in the decoder-only case we must carve a
source-restricted row out of full-prompt self-attention and explicitly account
for the residual mass on the accepted target prefix, the rest of the prompt,
and the speculative suffix. Eq.~\eqref{eq:stop-frontier} is therefore the
decoder-only analogue of the original ``do not emit if the token aligns to the
last $f$ source frames'' condition, with $N^{(k)}_{\mathrm{acc}} + b$ replacing
the forbidden source tail.

\paragraph{Alignment heads.}
Not all $LH$ self-attention heads carry alignment signal; many perform
syntactic or coreferential roles. We posit the existence of a small
head subset $\mathcal{H} \subset \{1,\ldots,L\} \times \{1,\ldots,H\}$,
$|\mathcal{H}| \ll LH$, such that the head-averaged source row
$\frac{1}{|\mathcal{H}|} \sum_{(\ell,h) \in \mathcal{H}}
\mathbf{A}^{(\ell,h)}_{t, \phi^{(k)}(s)}$
approximates a gold source-to-target alignment on a held-out calibration set
$\mathcal{G} = \{(\mathbf{s}^{(i)}, \mathbf{y}^{(i)}, \mathbf{a}^{(i)})\}_i$,
where $\mathbf{a}^{(i)} \in \{1,\ldots,|\mathbf{s}^{(i)}|\}^{|\mathbf{y}^{(i)}|}$
is a forced word-level alignment. The sparsity assumption is consistent with
recent evidence that multilingual LLMs contain a small set of
translation-specific token alignment heads \cite{tokenheads26}, although our
selection is calibrated directly on the prompt layout and task role used at
inference time.

\paragraph{Offline head selection.}
For a single $(\ell, h)$, define the argmax alignment
\begin{equation}
  \hat{a}^{(\ell,h)}_t(\mathbf{s}, \mathbf{y})
  = \arg\max_{s} \mathbf{A}^{(\ell,h)}_{t, \phi(s)},
  \label{eq:head-argmax}
\end{equation}
and its token-level alignment accuracy on $\mathcal{G}$,
\begin{equation}
  \mathrm{Acc}_{\mathrm{align}}(\ell, h)
  = \frac{1}{|\mathcal{G}|}
  \sum_i \frac{1}{|\mathbf{y}^{(i)}|}
  \sum_{t=1}^{|\mathbf{y}^{(i)}|}
  \mathbf{1}\!\bigl[\hat{a}^{(\ell,h)}_t = \mathbf{a}^{(i)}_t\bigr],
  \label{eq:align-acc}
\end{equation}
where $\hat{a}^{(\ell,h)}_t$ is evaluated at $(\mathbf{s}^{(i)}, \mathbf{y}^{(i)})$.
We then select $\mathcal{H} = \operatorname*{top-}k_{(\ell,h)} \mathrm{Acc}_{\mathrm{align}}(\ell, h)$
with $k = 8$ throughout. The calibration corpus $\mathcal{G}$ is built from
forced-alignment outputs on the dev set (MuST-C + in-house).

Figure~\ref{fig:decoder-only-heads} gives the architecture-aware view used in
practice for Gemma-4: the same layer$\times$head score map is overlaid with
the full-attention bands, the late shared-KV region, and a reminder that the
grid indexes query heads whereas capture cost still follows KV ownership.
\input{generated/decoder_only_alignatt_heads.tex}

\paragraph{One abstraction, two roles.}
The construction of Eqs.~\eqref{eq:layout}--\eqref{eq:align-acc} is
\emph{role-agnostic}. Let $r \in \{\mathrm{ASR}, \mathrm{MT}\}$ index the
role: for $r = \mathrm{ASR}$, $\mathbf{s}$ is a sequence of audio-token
embeddings and $\mathbf{y}$ the transcript; for $r = \mathrm{MT}$,
$\mathbf{s}$ is the transcript and $\mathbf{y}$ the translation. The same
LLM weights are used in both roles; only the prompt layout and the head set
$\mathcal{H}$ depend on $r$. We write $\mathcal{H}_r$ for the
role-specific head set, and observe empirically that
$\bigl|\mathcal{H}_{\mathrm{ASR}} \cap \mathcal{H}_{\mathrm{MT}}\bigr| \ll
\min(|\mathcal{H}_{\mathrm{ASR}}|, |\mathcal{H}_{\mathrm{MT}}|)$: the heads that
carry audio-to-text alignment are largely distinct from those that carry
source-to-target alignment, consistent with per-role functional specialization. The abstraction is the same in both roles; what
remains is to realize it operationally. Sec.~\ref{sec:qk-fast} shows how to
reconstruct only the draft-to-source slice needed by the policy, and how to
obtain the required query/key tensors under graph-compiled inference where
the native attention matrix is never materialized.

% ============================================================
% SECTION 4 — METHOD
% ============================================================
\section{Selective Reconstruction and Engine-Native Capture}
\label{sec:qk-fast}
\label{sec:vllm}

\paragraph{Method overview.}
Section~\ref{sec:alignatt-llm} defined the object we want: source-restricted
self-attention on a small set of alignment heads. The method has two coupled
parts. First, we reconstruct exactly only the draft-to-source attention block
that the acceptance policy consumes. Second, we make the required query/key
tensors observable inside a fused, graph-compiled runtime where the full
attention matrix never appears in memory.

\subsection{Selective Reconstruction for Acceptance}

\paragraph{Per-head attention and the policy-visible block.}
At decoding step $k$, let $\mathbf{p}^{(k)}$ be the prompt of
Eq.~\eqref{eq:layout}, with local source index set
$\mathcal{S}^{(k)} = \mathcal{S}^{(k)}_{\mathrm{acc}} \sqcup
\mathcal{S}^{(k)}_{\mathrm{inacc}}$, non-source context
$\mathcal{P} = \mathcal{P}_1 \cup \mathcal{P}_2$, and draft
$\mathcal{D}$, and prompt-position map
$\phi^{(k)} : \mathcal{S}^{(k)} \hookrightarrow \{1,\ldots,n\}$. Here
$\mathcal{S}^{(k)}_{\mathrm{acc}}$ and
$\mathcal{S}^{(k)}_{\mathrm{inacc}}$ are the accessible and inaccessible parts
of the live source at the current decode step. For each attention layer
$\ell \in \{1,\ldots,L\}$ and head $h \in \{1,\ldots,H\}$, let
$\mathbf{Q}^{(\ell,h)}, \mathbf{K}^{(\ell,h)}
\in \mathbb{R}^{n \times d_\ell}$ be the post-RoPE, post-norm query and key
tensors. The single-head attention row at draft position $t \in \mathcal{D}$
is
\begin{equation}
  \mathbf{A}^{(\ell,h)}_{t,\cdot}
  = \mathrm{softmax}\!\left(
  \mathbf{Q}^{(\ell,h)}_t \bigl(\mathbf{K}^{(\ell,h)}\bigr)^{\!\top}
  + \mathbf{m}_t
  \right),
  \label{eq:softmax}
\end{equation}
with $\mathbf{m}_t$ the causal (optionally sliding-window) mask. Gemma-4
folds the usual $1/\sqrt{d}$ scaling into $\mathbf{Q}$. Fused kernels compute
$\mathbf{A}^{(\ell,h)} \mathbf{V}^{(\ell,h)}$ without materializing
$\mathbf{A}^{(\ell,h)}$.

\paragraph{\emph{qk-fast} selective reconstruction.}
Let $\mathcal{H} \subset \bigcup_\ell \{\ell\} \times \{1,\ldots,H\}$ be the
AlignAtt head set, with $|\mathcal{H}| \ll LH$, selected offline from the
alignment-head calibration corpus. The acceptance policy never consumes the
full $n \times n$ matrix: it only needs rows indexed by the draft and columns
indexed by the source span. We therefore reconstruct only that block:
\begin{equation}
  \widehat{\mathbf{A}}^{(\ell,h)}_{t,s}
  = \frac{\exp\!\bigl(\mathbf{Q}^{(\ell,h)}_t \!\cdot\! \mathbf{K}^{(\ell,h)}_{\phi^{(k)}(s)}\bigr)}
       {\sum_{j=1}^{n} \exp\!\bigl(\mathbf{Q}^{(\ell,h)}_t \!\cdot\! \mathbf{K}^{(\ell,h)}_j + m_{t,j}\bigr)},
  \label{eq:qk-fast}
\end{equation}
for $t \in \mathcal{D}$, $s \in \mathcal{S}^{(k)}$, and $(\ell,h) \in \mathcal{H}$.
Eq.~\eqref{eq:qk-fast} recovers \emph{exactly} the same source-reindexed
softmax block as Eq.~\eqref{eq:softmax}; the only discrepancy is floating-point
reassociation between the fused kernel and our out-of-band recomputation. The
denominator still ranges over all $n$ prompt positions, so the runtime must
expose $\mathbf{K}^{(\ell,h)}$ for the full prompt even though the reconstructed
object is indexed only by source positions $s \in \mathcal{S}^{(k)}$.
Figure~\ref{fig:qk-fast} illustrates the construction.

\begin{figure*}[t]
\centering
\resizebox{\linewidth}{!}{%
\begin{tikzpicture}[
  x=1cm, y=1cm,
  >={Latex[length=2.6mm]},
  every node/.style={font=\small},
  panel/.style={draw=black!45, rounded corners=3pt, line width=0.6pt, fill=white},
  title/.style={font=\bfseries\small, text=black!85},
  subtitle/.style={font=\scriptsize, text=black!60, align=left},
  stagearrow/.style={->, line width=1.1pt, draw=black!75},
  guide/.style={draw=black!25, line width=0.35pt, dashed},
  kernelbox/.style={draw=black!35, dashed, fill=black!3, line width=0.6pt},
  captureband/.style={fill=orange!20, draw=orange!75!black, line width=0.9pt},
  captureline/.style={draw=orange!75!black, line width=0.9pt, rounded corners=1pt},
  reconbox/.style={draw=teal!65!black, fill=teal!15, line width=1.1pt, rounded corners=1.5pt},
  partP/.style={draw=black!30, fill=black!7, line width=0.3pt},
  partPy/.style={draw=green!35!black, fill=green!18, line width=0.3pt},
  partSa/.style={draw=blue!40, fill=blue!18, line width=0.3pt},
  partSi/.style={draw=blue!25, fill=blue!6, line width=0.3pt},
  partD/.style={draw=purple!45, fill=purple!14, line width=0.3pt},
  notrecon/.style={pattern=north east lines, pattern color=black!25, draw=black!20, line width=0.2pt},
  labelnote/.style={font=\scriptsize, text=black!70, align=center},
  mathnote/.style={font=\scriptsize\itshape, text=black!60, align=center},
]

  %% ===== Panel 1: fused forward with capture sites =====
  \draw[panel] (-1.55, -0.4) rectangle (9.3, 7.3);
  \node[title, anchor=north west] at (-1.30, 7.18) {1.~Fused attention forward};

  % matrix geometry
  \def\mL{1.70}\def\mR{7.70}\def\mB{0.55}\def\mT{5.55}
  % column boundaries (keys): P_1 | S_acc | S_inacc | P_2-instr | P_2-committed | D
  \def\cPa{2.10}   % P_1     : 1.70-2.10  (short system prompt)
  \def\cSa{3.70}   % S_acc   : 2.10-3.70  (visible accessible source)
  \def\cSi{4.60}   % S_inacc : 3.70-4.60  (visible inaccessible source)
  \def\cPi{5.30}  % instr   : 4.60-5.30
  \def\cPy{6.90}  % y_{1:m_k}: 5.30-6.90 (already committed target prefix - GROWS across steps)
  % D : 6.90-7.70 (narrow draft)
  % row boundaries (top to bottom order): P_1 | S_acc | S_inacc | P_2-instr | P_2-committed | D
  \def\rPa{4.95}   % P_1     : 4.95-5.55
  \def\rSa{3.75}   % S_acc   : 3.75-4.95
  \def\rSi{2.95}   % S_inacc : 2.95-3.75
  \def\rPi{2.40}  % instr   : 2.40-2.95
  \def\rPy{1.25}  % y_{1:m_k}: 1.25-2.40
  % D : 0.55-1.25 (narrow)

  % matrix body (dashed -> never stored)
  \draw[kernelbox] (\mL, \mB) rectangle (\mR, \mT);
  \node[mathnote] at ({(\mL+\mR)/2}, 4.40)
    {$\mathbf{A}^{(\ell,h)}=\mathrm{softmax}\!\bigl(\mathbf{Q}\mathbf{K}^{\!\top}+\mathbf{m}\bigr)$};
  \node[labelnote] at ({(\mL+\mR)/2}, 3.95) {$n{\times}n$, fused, never materialized};

  % internal partition guides
  \foreach \x in {\cPa,\cSa,\cSi,\cPi,\cPy} \draw[guide] (\x,\mB) -- (\x,\mT);
  \foreach \y in {\rPa,\rSa,\rSi,\rPi,\rPy} \draw[guide] (\mL,\y) -- (\mR,\y);

  % --- Hatch "not reconstructed" on D rows outside S^{(k)} columns ---
  \draw[notrecon] (\mL,  \mB) rectangle (\cPa,  \rPy);  % D x P_1
  \draw[notrecon] (\cSi, \mB) rectangle (\mR,   \rPy);  % D x (P_2 u D)

  % target block: D rows x (S_acc U S_inacc) cols
  \draw[reconbox] (\cPa, \mB) rectangle (\cSi, \rPy);
  \node[font=\scriptsize\bfseries, text=teal!55!black]
    at ({(\cPa+\cSi)/2}, {(\mB+\rPy)/2 + 0.07})
    {$\widehat{\mathbf{A}}_{\mathcal{D},\mathcal{S}^{(k)}}$};
  \node[font=\tiny, text=teal!55!black]
    at ({(\cPa+\cSi)/2}, {(\mB+\rPy)/2 - 0.17})
    {$L_k{\times}N_k$, only block used};

  % --- column partition strip (keys) ---
  \def\csB{5.68}\def\csT{6.00}
  \draw[partP ] (\mL  ,\csB) rectangle (\cPa ,\csT) node[midway,font=\tiny]{$\mathcal{P}_1$};
  \draw[partSa] (\cPa ,\csB) rectangle (\cSa ,\csT) node[midway,font=\tiny]{$\mathcal{S}^{(k)}_{\mathrm{acc}}$};
  \draw[partSi] (\cSa ,\csB) rectangle (\cSi ,\csT) node[midway,font=\tiny]{$\mathcal{S}^{(k)}_{\mathrm{inacc}}$};
  \draw[partP ] (\cSi ,\csB) rectangle (\cPi,\csT) node[midway,font=\tiny]{instr};
  \draw[partPy] (\cPi,\csB) rectangle (\cPy,\csT) node[midway,font=\tiny,text=green!30!black]{$\mathbf{y}_{1:m_k}$};
  \draw[partD ] (\cPy,\csB) rectangle (\mR  ,\csT) node[midway,font=\tiny]{$\mathcal{D}$};
  % P_2 bracket under instr|committed columns
  \draw[black!50, line width=0.3pt] (\cSi, \csB-0.05) -- (\cSi, \csB-0.14) -- (\cPy, \csB-0.14) -- (\cPy, \csB-0.05);
  \node[labelnote, anchor=north, font=\tiny] at ({(\cSi+\cPy)/2}, \csB-0.15) {$\mathcal{P}_2$};
  % orange capture bracket above keys
  \draw[captureline] ({\mL-0.05}, {\csT+0.14}) -- ({\mR+0.05}, {\csT+0.14});
  \draw[captureline] ({\mL-0.05}, {\csT+0.06}) -- ({\mL-0.05}, {\csT+0.14});
  \draw[captureline] ({\mR+0.05}, {\csT+0.06}) -- ({\mR+0.05}, {\csT+0.14});
  \node[labelnote, anchor=south] at ({(\mL+\mR)/2}, {\csT+0.18})
    {key columns in the full prompt\\[-1pt]
     \textcolor{orange!70!black}{\textbf{capture all} $\mathbf{K}^{(\ell,h)}$ columns}};

  % --- row partition strip (queries) ---
  \def\rsL{1.20}\def\rsR{1.52}
  \draw[partP ] (\rsL,\rPa ) rectangle (\rsR,\mT  ) node[midway,rotate=90,font=\tiny]{$\mathcal{P}_1$};
  \draw[partSa] (\rsL,\rSa ) rectangle (\rsR,\rPa ) node[midway,rotate=90,font=\tiny]{$\mathcal{S}^{(k)}_{\mathrm{acc}}$};
  \draw[partSi] (\rsL,\rSi ) rectangle (\rsR,\rSa ) node[midway,rotate=90,font=\tiny]{$\mathcal{S}^{(k)}_{\mathrm{inacc}}$};
  \draw[partP ] (\rsL,\rPi) rectangle (\rsR,\rSi ) node[midway,rotate=90,font=\tiny]{instr};
  \draw[partPy] (\rsL,\rPy) rectangle (\rsR,\rPi) node[midway,rotate=90,font=\tiny,text=green!30!black]{$\mathbf{y}_{1:m_k}$};
  \draw[partD ] (\rsL,\mB  ) rectangle (\rsR,\rPy) node[midway,rotate=90,font=\tiny]{$\mathcal{D}$};
  \node[labelnote, rotate=90] at ({\rsL-0.58}, {(\mB+\mT)/2})
    {query rows in the full prompt};

  % orange capture bracket around the D row slice only
  \draw[captureline] ({\rsL-0.14}, \mB) -- ({\rsL-0.14}, \rPy);
  \draw[captureline] ({\rsL-0.14}, \mB) -- ({\rsL-0.06}, \mB);
  \draw[captureline] ({\rsL-0.14}, \rPy) -- ({\rsL-0.06}, \rPy);
  \node[labelnote, text=orange!70!black, anchor=east, align=right, text width=1.55cm]
    at (0.30, {(\mB+\rPy)/2})
    {\textbf{capture only}\\ $\mathbf{Q}^{(\ell,h)}_{\mathcal{D}}$\\(draft rows)};

  %% ===== Panel 2: selective reconstruction =====
  \draw[panel] (10.3, -0.4) rectangle (15.9, 7.3);
  \node[title, anchor=north west] at (10.55, 7.18) {2.~Recompute Only That Block};
  \node[subtitle, anchor=north west, text width=5.05cm] at (10.55, 6.72)
    {Exact \emph{qk-fast} recomputation from captured $Q/K$; then average over $\mathcal{H}$, split accessible vs.\ inaccessible source mass, and apply the acceptance gates.};

  % block geometry
  \def\bL{11.25}\def\bR{15.45}\def\bMid{13.48}
  \def\bB{0.95}\def\bT{3.15}
  \def\bcB{3.28}\def\bcT{3.60}

  % column partition strip (S_acc | S_inacc) above block
  \draw[partSa] (\bL,\bcB) rectangle (\bMid,\bcT);
  \node[labelnote] at ({(\bL+\bMid)/2}, {(\bcB+\bcT)/2}) {$\mathcal{S}^{(k)}_{\mathrm{acc}}$};
  \draw[partSi] (\bMid,\bcB) rectangle (\bR,\bcT);
  \node[labelnote] at ({(\bMid+\bR)/2}, {(\bcB+\bcT)/2}) {$\mathcal{S}^{(k)}_{\mathrm{inacc}}$};
  \node[labelnote, anchor=south] at ({(\bL+\bR)/2}, {\bcT+0.04})
    {source columns in local live-source order};

  % row label D
  \def\brL{10.92}\def\brR{11.18}
  \draw[partD] (\brL,\bB) rectangle (\brR,\bT);
  \node[labelnote, rotate=90] at ({(\brL+\brR)/2}, {(\bB+\bT)/2}) {$\mathcal{D}$};
  \node[labelnote, rotate=90] at (10.64, {(\bB+\bT)/2})
    {draft query rows};

  % block
  \draw[reconbox] (\bL,\bB) rectangle (\bR,\bT);
  \draw[teal!55!black, dashed, line width=0.4pt] (\bMid,\bB) -- (\bMid, 1.68);
  \draw[teal!55!black, dashed, line width=0.4pt] (\bMid, 2.42) -- (\bMid,\bT);
  \node[font=\small, text=teal!55!black]
    at ({(\bL+\bR)/2}, {(\bB+\bT)/2 + 0.18})
    {$\widehat{\mathbf{A}}^{(\ell,h)}_{\mathcal{D},\,\mathcal{S}^{(k)}}$};
  \node[labelnote, text=teal!55!black]
    at ({(\bL+\bR)/2}, {(\bB+\bT)/2 - 0.20})
    {$L_k$ draft rows $\times$ $N_k$ source columns};

  %% ===== arrow between panels (horizontal, no diagonal) =====
  \draw[stagearrow]
    (9.35, {(\mB+\rPy)/2})
    -- (11.08, {(\mB+\rPy)/2});
\end{tikzpicture}
}%
\caption{\textbf{Selective reconstruction with engine-native capture.} The
$n{\times}n$ attention matrix
$\mathbf{A}^{(\ell,h)}$ is executed entirely inside the fused attention
kernel, so the full $n{\times}n$ matrix is never materialized. For each
AlignAtt head $(\ell,h)\in\mathcal{H}$, the observer copies
$\mathbf{K}^{(\ell,h)}$ for all positions and $\mathbf{Q}^{(\ell,h)}$ only for
the draft rows $\mathcal{D}$ into fixed-shape buffers. The green band
$\mathbf{y}_{1:m_k}$ inside $\mathcal{P}_2$ marks the committed target prefix
that grows across chunks; hatched cells are never reconstructed. After the
forward, Eq.~\eqref{eq:qk-fast} recomputes exactly only the draft-to-source
block $\widehat{\mathbf{A}}_{\mathcal{D},\mathcal{S}^{(k)}}$, which is the sole
object consumed by the source-side provenance scores of
Eqs.~\eqref{eq:prov-acc}--\eqref{eq:provenance} and by the
acceptance policy of this section.}
\label{fig:qk-fast}
\end{figure*}

\paragraph{From reconstructed rows to provenance features.}
The averaged source row for token $t$ is
\begin{equation}
  \bar{\mathbf{r}}_t(s) \;=\; \frac{1}{|\mathcal{H}|} \sum_{(\ell,h) \in \mathcal{H}}
  \widehat{\mathbf{A}}^{(\ell,h)}_{t,s},
  \qquad s \in \mathcal{S}^{(k)},
  \label{eq:rowbar}
\end{equation}
and the corresponding source-side provenance scores are
\begin{align}
  \pi_t^{\mathrm{acc}}
  &\;=\;
  \sum_{s \in \mathcal{S}^{(k)}_{\mathrm{acc}}} \bar{\mathbf{r}}_t(s),
  \label{eq:prov-acc}\\
  \pi_t^{\mathrm{inacc}}
  &\;=\;
  \sum_{s \in \mathcal{S}^{(k)}_{\mathrm{inacc}}} \bar{\mathbf{r}}_t(s).
  \label{eq:provenance}
\end{align}
These two quantities partition the total source mass according to the
accessibility frontier. Residual mass on non-source prompt positions and on the
draft is not used by the acceptance rule and is recovered only in the full
prompt-space replay of Sec.~\ref{sec:vllm} when diagnostic accounting is
needed.
The raw averaged row $\bar{\mathbf{r}}_t$ is useful as a source-mass feature,
but the source argmax itself is more stable when computed from a
per-head normalized representation. Let
$\mathbf{r}_t \in \mathbb{R}^{|\mathcal{H}| \times N_k}$ stack the
per-head source rows,
\[
  [\mathbf{r}_t]_{(\ell,h),\, s}
  \;=\;
  \widehat{\mathbf{A}}^{(\ell,h)}_{t,s}.
\]

\paragraph{Prefix-online Welford normalization.}
To make the per-token argmax robust to head-scale drift along the draft, we
normalize each row against the running moments
$(\boldsymbol{\mu}_t, \boldsymbol{\sigma}^2_t)$ of all prior rows,
maintained by a standard prefix-online Welford recurrence over
$\mathbf{r}_t \in \mathbb{R}^{|\mathcal{H}| \times N_k}$:
\begin{equation}
  \tilde{\mathbf{r}}_t \;=\;
  \bigl(\mathbf{r}_t - \boldsymbol{\mu}_t\bigr) \;\oslash\;
  \sqrt{\boldsymbol{\sigma}^2_t + \varepsilon},
  \qquad \varepsilon = 10^{-6}.
  \label{eq:welford-norm}
\end{equation}
We then apply a 1-D median filter of width $w$ independently to each head along
the source axis,
\begin{equation}
  u_t^{(\ell,h)}(s) =
  \mathrm{median}_{|j| \le (w-1)/2}\, \tilde{r}_t^{(\ell,h)}(s+j),
  \label{eq:medfilt}
\end{equation}
with default $w = 7$. The aligned source position is then
\begin{equation}
  \hat{s}_t = \arg\max_{s} \bar{u}_t(s),
  \quad
  \bar{u}_t(s) = \tfrac{1}{|\mathcal{H}|} \sum_{(\ell,h)} u_t^{(\ell,h)}(s),
  \label{eq:argmax}
\end{equation}
over $s \in \{1,\ldots,N_k\}$.

\paragraph{Acceptance policy.}
The policy $\Pi$ of Eq.~\eqref{eq:accept-prefix} is a first-failure scan
over the draft indices $t = 1, 2, \ldots, L_k$ with four stop conditions.
Let $\hat{s}_{t-1}$ denote the previous accepted alignment (undefined at
$t=1$), let $\rho \in \mathbb{N}$ be the rewind threshold
(default $3$), and $b \in \mathbb{N}$ the border margin (default $0$):
\begin{align}
  \text{\textsc{rewind}} :\quad
  &\hat{s}_{t-1} - \hat{s}_t \;>\; \rho,
  \label{eq:stop-rewind}\\
  \text{\textsc{source-frontier}} :\quad
  &\hat{s}_t \;\geq\; N^{(k)}_{\mathrm{acc}} + b,
  \label{eq:stop-frontier}\\
  \text{\textsc{argmax-mass-weak}} :\quad
  &\bar{\mathbf{r}}_t(\hat{s}_t) \;<\; \tau_{\mathrm{argmax}},
  \label{eq:stop-argmax}\\
  \text{\textsc{provenance-weak}} :\quad
  &\pi_t^{\mathrm{acc}} \;<\; \tau_{\mathrm{src}},
  \label{eq:stop-prov}
\end{align}
all evaluated in order. The accepted length is the largest
$\ell_k$ such that none of
Eq.~\eqref{eq:stop-rewind}--\eqref{eq:stop-prov} fires for any
$t \leq \ell_k$. Finally, $\ell_k$ is rounded back to the nearest
stability-unit boundary (end of a whitespace- or punctuation-delimited
word), so partial subword fragments are never emitted.

\paragraph{Monotone source-side gating under frontier refinement.}
We isolate the precise part of the policy that is monotone in the accessibility
frontier.

\begin{proposition}[Frontier-monotone source-side gates]
\label{prop:monotone}
Fix a draft token $t$ and its reconstructed source row
$\bar{\mathbf{r}}_t$ on a fixed live source index set $\mathcal{S}^{(k)}$.
Suppose the accessibility partition is refined by moving source indices from
$\mathcal{S}^{(k)}_{\mathrm{inacc}}$ to a larger accessible set
$\mathcal{S}'_{\mathrm{acc}} \supseteq \mathcal{S}^{(k)}_{\mathrm{acc}}$, with
$N'_{\mathrm{acc}} := |\mathcal{S}'_{\mathrm{acc}}|$, while leaving
$\bar{\mathbf{r}}_t$ unchanged. If
$\hat{s}_t < N^{(k)}_{\mathrm{acc}} + b$
(Eq.~\eqref{eq:stop-frontier}) and
$\pi^{\mathrm{acc}}_t \geq \tau_{\mathrm{src}}$
(Eq.~\eqref{eq:stop-prov}) before the refinement, then both inequalities remain
satisfied after the refinement.
\end{proposition}
\begin{proof}[Proof sketch]
The frontier gate Eq.~\eqref{eq:stop-frontier} compares $\hat{s}_t$ against a
threshold that increases from $N^{(k)}_{\mathrm{acc}} + b$ to
$N'_{\mathrm{acc}} + b$, so once the inequality is satisfied it remains
satisfied after the refinement. The provenance gate
Eq.~\eqref{eq:stop-prov} depends on the same reconstructed attention row and on
the accessible set through Eq.~\eqref{eq:prov-acc}; as
$\mathcal{S}^{(k)}_{\mathrm{acc}}$ grows to $\mathcal{S}'_{\mathrm{acc}}$,
$\pi^{\mathrm{acc}}_t$ is
non-decreasing. Both source-side gates are therefore monotone under frontier
refinement.
\qedhere
\end{proof}

Proposition~\ref{prop:monotone} is what makes the monotone emission rule of
Eq.~\eqref{eq:monotone} coherent at the policy level: relabeling source
positions from inaccessible to accessible cannot, by itself, flip an already
accepted token back to rejected. Any later change in decision must come from a
different draft token or a different reconstructed attention row, not from the
frontier split alone.

\paragraph{Robustness to reconstruction error.}
Write $\mathbf{A}^{(\ell,h)}_{t,s} :=
\mathbf{A}^{(\ell,h)}_{t,\phi^{(k)}(s)}$ for the exact source-reindexed block.
Let $\varepsilon_\infty = \max_{(\ell,h,t,s)} \bigl|
  \widehat{\mathbf{A}}^{(\ell,h)}_{t,s} - \mathbf{A}^{(\ell,h)}_{t,s}\bigr|$
be the worst-case reconstruction error due to fp-reassociation. Under the
median filter of width $w$ and Welford normalization, a sufficient condition
for the argmax of Eq.~\eqref{eq:argmax} to be invariant under replacing
$\widehat{\mathbf{A}}$ by $\mathbf{A}$ is
\begin{equation}
  \varepsilon_\infty \;<\; \tfrac{1}{2} \bigl(\bar{u}_t(\hat{s}_t)
  - \max_{s \neq \hat{s}_t} \bar{u}_t(s)\bigr),
  \label{eq:robust}
\end{equation}
i.e.\ the argmax margin dominates the reconstruction perturbation. Empirically
$\varepsilon_\infty \leq 1.2 \times 10^{-2}$ on Gemma-4 E4B prompts, which is
well below typical argmax margins under Welford-normalised median filtering;
accordingly we observe identical policy decisions at every sampled step.

\paragraph{Complexity.}
Materializing $\mathbf{A}$ over all layers and heads costs
$\Theta(LHn^2)$ per forward.
Selective reconstruction costs, per draft step,
\begin{equation}
  \mathcal{C}_{\text{qk-fast}}
  \;=\;
  \Theta\!\bigl(|\mathcal{H}| \cdot |\mathcal{D}| \cdot n \cdot \bar{d}\bigr)
  \;+\;
  \Theta\!\bigl(|\mathcal{H}| \cdot |\mathcal{D}| \cdot N_k \cdot w\bigr),
  \label{eq:complexity}
\end{equation}
with $\bar{d} \in \{256, 512\}$ on Gemma-4 and the second term accounting for
the median filter. For $|\mathcal{H}|{=}8$, $|\mathcal{D}|{\leq}16$,
$n{\leq}1024$, $N_k{\leq}128$, $w{=}7$, the total is
$< 6 \times 10^5$ scalar multiplications per decoding step.

\subsection{Engine-Native Realization under Compiled Inference}

\paragraph{Why runtime observability is part of the method.}
Eq.~\eqref{eq:qk-fast} is only useful if the runtime exposes the selected
query and key tensors. Under fused attention kernels (FlashAttention family)
and ahead-of-time graph compilation (\texttt{cudagraph=full}), it does not:
the Q/K tensors never leave the fused kernel, Python-level recorders cannot
hook into the compiled graph, and unused operator outputs are removed by
dead-code elimination during the \texttt{torch.inductor} lowering pass. The
observer below is therefore not ancillary tooling; it is the mechanism that
makes selective reconstruction available in the actual inference stack.

\paragraph{Capture primitive.}
We install at import time a per-layer stateless observer with three
fixed-shape buffers: one for prompt keys, one for decode queries, and one for
decode keys. Let $\mathcal{H}_\ell = \{h : (\ell,h) \in \mathcal{H}\}$.
Capture is a single static scatter-add per buffer,
\begin{equation}
  \mathbf{B}^{(\ell)}_{\star} \mathrel{+\!\!=} \mathrm{scatter}\!\left(
    \mathrm{pos}_{\star}(\ell),\, \mathbf{X}^{(\ell, \mathcal{H}_\ell)}_{\star}
  \right),
  \label{eq:scatter}
\end{equation}
for $\mathbf{X}_\star \in \{\mathbf{Q}^{(\ell)}, \mathbf{K}^{(\ell)}\}$,
implemented as \texttt{torch.Tensor.scatter\_add\_} over statically-shaped
index tensors. Fixed-shape scatter-adds are compile-safe: they avoid
\texttt{nonzero()} and Python conditionals. The captured graph therefore stays
independent of runtime shapes within one \texttt{cudagraph=full} bucket.

\paragraph{Compile-visible custom op.}
Let $\Phi : (\ell, \mathrm{pos}, \mathbf{Q}, \mathbf{K}) \mapsto \mathbf{0}$
be the capture operator, registered as
$\mathtt{alignatt{::}capture\_mt\_qk}$ with declared purity
$\mathtt{mutates\_args=()}$. This makes \texttt{torch.compile} treat $\Phi$
as a value-returning operator whose
writes to the observer buffers remain outside the alias analyser. This is what
allows $\Phi$ to be fused into the compiled graph rather than rejected as a
side-effecting barrier.

\paragraph{Sentinel dependency against dead-code elimination.}
Under \texttt{cudagraph=full}, any call whose return value does not flow into
the graph output is eliminated. To prevent elimination without altering
numerics, the patched attention forward injects a zero sentinel:
\begin{equation}
  \mathbf{o}'_\ell = \mathbf{o}_\ell + \Phi\bigl(\ell, \mathrm{pos}, \mathbf{Q}^{(\ell)}, \mathbf{K}^{(\ell)}\bigr),
  \label{eq:sentinel}
\end{equation}
with $\mathbf{o}_\ell = \mathbf{A}^{(\ell)} \mathbf{V}^{(\ell)}$ and $\Phi \equiv \mathbf{0}$,
which is numerically identical to $\mathbf{o}_\ell$ but places $\Phi$ in the
transitive fan-in of the graph output, so $\Phi$ cannot be DCE-pruned. This
is the smallest algebraic perturbation sufficient to keep the observer alive
under graph compilation.

\paragraph{Replay consistency.}
AOT-compiled graphs resolve attributes by direct lookup. To avoid a first-pass
attribute miss after process restart, every
per-layer observer slot $\mathtt{O}_\ell$ is pre-seeded at patch-install
time. If layer $\ell$ has no live observer, $\Phi$ returns immediately. The
capture path is thus
gated purely by whether the backend has installed a live observer, without
any graph recompilation.

\paragraph{Offline reconstruction.}
At the end of each decode step we read the buffers and replay the draft rows in
prompt space, with prompt keys and draft keys stored separately.
Writing $\mathbf{B}^{(\ell,h)}_\star := \mathbf{B}^{(\ell)}_\star[h]$ for brevity,
\begin{align}
  \mathbf{P}^{(\ell,h)} &=
  \mathbf{B}^{(\ell,h)}_{\mathrm{dQ}}
  \bigl(\mathbf{B}^{(\ell,h)}_{\mathrm{pK}}\bigr)^{\!\top},
  \label{eq:prompt-logits}\\
  \mathbf{S}^{(\ell,h)} &=
  \mathbf{B}^{(\ell,h)}_{\mathrm{dQ}}
  \bigl(\mathbf{B}^{(\ell,h)}_{\mathrm{dK}}\bigr)^{\!\top}
  + \mathbf{m}_{\mathrm{causal}},
  \label{eq:suffix-logits}\\
  \widetilde{\mathbf{A}}^{(\ell,h)}
  &= \mathrm{softmax}[\mathbf{P}^{(\ell,h)}, \mathbf{S}^{(\ell,h)}],
  \label{eq:vllm-softmax}
\end{align}
with $\mathbf{m}_{\mathrm{causal}}$ the strict upper-triangular $-\infty$
mask. Restricting $\widetilde{\mathbf{A}}^{(\ell,h)}$ to the source columns
$\phi^{(k)}(s)$ and reindexing by $s \in \mathcal{S}^{(k)}$ recovers the block
of Eq.~\eqref{eq:qk-fast}, and hence $\bar{\mathbf{r}}_t$ and the source-side
provenance scores of Eqs.~\eqref{eq:rowbar}--\eqref{eq:provenance}. The
remaining columns yield the residual mass on $\mathcal{P}$ and $\mathcal{D}$
when diagnostic accounting is needed.

\paragraph{Qualitative reconstruction example.}
Figure~\ref{fig:mt-selective-reconstruction} instantiates the full
reconstruction stack on a text-only MT probe. We display the exact prompt
partition seen by the decoder, aggregate reconstructed token-level rows to the
word level for readability, and keep the four-way provenance accounting visible
for every drafted word. The selected snapshot is taken immediately before the
AlignAtt gate rejects the next word, so the figure shows both alignment quality
and the operational decision boundary.
\input{generated/mt_selective_reconstruction_example.tex}

The green band in Fig.~\ref{fig:mt-selective-reconstruction} is the already
accepted target prefix $\mathbf{y}_{1:m_k}$ carried over from earlier
streaming steps; it is causal context for the decoder, not the object of the
frontier gate. The orange words are the current draft
$\widetilde{\mathbf{y}}^{(k)}$, scanned left-to-right by the policy at the
current step $k$. Thus $\hat{s}_{t-1}$ in Eq.~\eqref{eq:stop-rewind} refers to
the previous drafted word \emph{within the same live draft}, not to a token
accepted at an earlier streaming step. A drafted word may still attract large
self-attention mass on the accepted prefix and yet remain admissible as long as
its source-side signal stays inside the accessible frontier; this is exactly
why the residual provenance terms are kept explicit in the decoder-only
formulation.
\input{generated/qualitative_provenance_summary.tex}

\paragraph{Numerical parity.}
Let $\mathbf{A}^{(\mathrm{TF})}$ denote the reference attention tensor
produced by an uncompiled Transformers forward on the identical prompt.
We observe
\begin{equation}
  \|\Delta\mathbf{A}\|_\infty \leq 1.2\!\times\!10^{-2},\;
  \|\Delta\mathbf{A}\|_1/(nL_k) \leq 4\!\times\!10^{-4},
  \label{eq:parity}
\end{equation}
with $\Delta\mathbf{A} := \widetilde{\mathbf{A}} - \mathbf{A}^{(\mathrm{TF})}$,
on a $167 \times 12$ reference row set, with bit-identical acceptance
decisions $\Pi\!\left(\widetilde{\mathbf{y}}\right)_{\mathrm{vLLM}}
\;=\; \Pi\!\left(\widetilde{\mathbf{y}}\right)_{\mathrm{TF,\,eager}}$ across
a curated parity set of $\geq 128$ prompts.
Under Eq.~\eqref{eq:robust} the observed
$\|\cdot\|_\infty$ margin is well below typical post-Welford argmax gaps, so
the parity is not accidental but a direct consequence of the robustness
bound.

\paragraph{Paper-only seam benchmark.}
To isolate capture cost from full SimulStream variability, we benchmark the
same fixed text-only prompt suite under three MT seams: a minimal
Transformers eager reference, a Transformers \emph{qk-fast} reference, and the
engine-native vLLM \emph{qk-fast} path used by the runtime. Each seam runs in
its own subprocess, with one unmeasured warmup pass followed by repeated hot
runs on the same 16 prompts. Figure~\ref{fig:mt-capture-speed} reports the
median latency per generated token, while the generated JSON artifact keeps the
per-length-bin and per-prompt results for appendix tables.
\input{generated/mt_capture_speed_benchmark.tex}

\paragraph{End-to-end throughput.}
With $\Phi$ fused into the compiled graph, the cost of the observer is
dominated by the scatter-adds of Eq.~\eqref{eq:scatter} and the offline
$\mathbf{Q}\mathbf{K}^\top$ of
Eqs.~\eqref{eq:prompt-logits}--\eqref{eq:suffix-logits}. On the control
clip we measure an RTF penalty below $3\%$ relative to the baseline
\emph{graph-compiled} runtime.

% ============================================================
% SECTION 5 — EXPERIMENTS
% ============================================================
\section{Experiments}
\label{sec:experiments}

\paragraph{Setup.}
All experiments run on a single NVIDIA A40 with a graph-compiled vLLM
runtime (\texttt{cudagraph=full}) and Gemma-4 E4B-it in bfloat16.
The evaluation corpus is the 21-clip IWSLT~2026 in-house dev-set
(${\sim}2.1$\,h total, academic research talks), resegmented by
OmniSTEval~v0.1.7. We report BLEU and chrF against the dev reference,
XCOMET-XL~\cite{xcometxl23} for translation adequacy, and the
compute-unaware and computation-aware variants of LongYAAL
\cite{longyaal25} for latency. For detailed ASR diagnostics we use a
single 360\,s clip (\texttt{ccpXHNfaoy}) for which word-level boundary
references are available. All cascades use synchronous ASR/MT under the
same chunked schedule.

\subsection{ASR Alignment: Qwen3 Forced vs Gemma AlignAtt}
\label{sec:exp-asr}

The single-substrate claim of Sec.~\ref{sec:alignatt-llm} rests on AlignAtt
being usable as the alignment backend in its own right, not only as the MT
policy. We therefore compare the Gemma AlignAtt backend against a dedicated
forced aligner (Qwen3-ASR~\cite{qwen3asr25} followed by the Qwen3
forced aligner) as the ASR component of the cascade, evaluated on the
\texttt{ccpXHNfaoy} clip with 47 reference word boundaries.

\begin{table}[t]
\centering
\small
\begin{tabular}{lcc}
\toprule
 & Qwen3 forced & Gemma AlignAtt \\
\midrule
WER $\downarrow$                 & $0.113$ & $0.269$ \\
CER $\downarrow$                 & $0.046$ & $0.122$ \\
Boundary F1 $\pm$3w $\uparrow$   & $0.878$ & $0.042$ \\
Mean $|$boundary lag$|$ (s) $\downarrow$ & $0.145$ & $0.403$ \\
Load time (s)                    & $63.7$  & $123.0$ \\
RTF (wall-clock) $\downarrow$    & $0.306$ & $0.628$ \\
\bottomrule
\end{tabular}
\caption{\textbf{ASR alignment comparison on \texttt{ccpXHNfaoy}
(360\,s).} Qwen3 forced uses a dedicated 3B-parameter aligner; Gemma
AlignAtt reuses the decoder-only LLM already needed for MT. Boundary F1
and lag are computed against 47 gold word boundaries with a
$\pm$3-word tolerance.}
\label{tab:asr-compare}
\end{table}

Table~\ref{tab:asr-compare} shows the expected direction of the
specialization gap: the forced aligner produces lower WER ($0.113$ vs
$0.269$), tighter boundaries ($0.145$\,s vs $0.403$\,s mean absolute
lag), and a higher boundary F1. AlignAtt's low boundary F1 reflects a
segmentation-granularity mismatch --- the streaming policy commits at
each decoder token rather than at prosodic boundaries --- and is
therefore not directly comparable without mapping both outputs to the
same unit. The key point for the present paper is that a single LLM
substrate produces alignment signals strong enough to drive streaming
commit decisions at all, at roughly twice the RTF of a dedicated
aligner, while removing the need to host a second 3B-parameter ASR model.

To complement the single-clip diagnostic, Table~\ref{tab:asr-21clip}
reports aggregate en--en (ASR-as-MT) metrics over the full 21-clip
dev-set at chunk size $800$\,ms. BLEU, chrF, and LongYAAL are corpus-level
scores from OmniSTEval; WER and CER are averaged across clips.

\begin{table}[t]
\centering
\small
\begin{tabular}{lcc}
\toprule
 & Qwen3 forced (LA) & Gemma AlignAtt \\
\midrule
Mean WER $\downarrow$           & $0.092$ & --- \\
Mean CER $\downarrow$           & $0.042$ & --- \\
BLEU $\uparrow$                 & $74.82$ & --- \\
chrF $\uparrow$                 & $90.81$ & --- \\
LongYAAL CU (ms) $\downarrow$   & $865.7$ & --- \\
LongYAAL CA (ms) $\downarrow$   & $672.7$ & --- \\
\bottomrule
\end{tabular}
\caption{\textbf{21-clip ASR alignment comparison}, en--en (ASR-as-MT),
chunk $800$\,ms. BLEU/chrF and LongYAAL are corpus-level on the
OmniSTEval-resegmented dev-set; WER/CER are per-clip means over the
21 clips. Gemma AlignAtt entries pending re-run.}
\label{tab:asr-21clip}
\end{table}

\subsection{Streaming Latency/Quality Frontier}
\label{sec:exp-frontier}

We next treat the same Gemma AlignAtt backend as an end-to-end
transcriber and sweep its two policy thresholds --- the frame threshold
$f$ (Eq.~\eqref{eq:stop-frontier}) and the rewind threshold $\rho$
(Eq.~\eqref{eq:stop-rewind}) --- to expose the streaming latency/quality
frontier on the full 21-clip dev-set under the en--en (ASR-as-MT) task.
Table~\ref{tab:asr-frontier} reports three operating points below
$1$\,s of CU-LongYAAL (the regime targeted by this work), plus a
quality-anchor point at $\rho{=}200$ that trades ${\sim}3.5$\,s of
CU for ${+}9$ BLEU.

\begin{table}[t]
\centering
\small
\begin{tabular}{l c c c c c}
\toprule
Config & $f$ & $\rho$ & BLEU $\uparrow$ & CU (ms) $\downarrow$ & CA (ms) $\downarrow$ \\
\midrule
A       & $4$ & $200$ & $34.32$ & $668$ & $363$ \\
B       & $4$ & $50$  & $35.44$ & $821$ & $368$ \\
C       & $5$ & $200$ & $35.03$ & $911$ & $612$ \\
\midrule
Q (quality anchor) & $4$ & $200$ & $44.35$ & $3561$ & $3359$ \\
\bottomrule
\end{tabular}
\caption{\textbf{Gemma AlignAtt en--en streaming frontier} on the
21-clip dev-set ($k{=}6$). Configurations A--C sit below one second of
CU-LongYAAL. Q uses the original (pre-fix) rewind logic at the same
$(f,\rho)$ as A and shows the upper quality bound reachable at
multi-second latency.}
\label{tab:asr-frontier}
\end{table}

Within the sub-1\,s regime the three configurations trace a narrow
band ($34.3$--$35.4$ BLEU; $668$--$911$\,ms CU), with $\rho{=}50$
(config B) offering the best BLEU and $f{=}4, \rho{=}200$ (config A)
the lowest CU. The ${\sim}9$ BLEU gap between A--C and the quality
anchor Q quantifies the cost of the sub-1\,s target: without the
anchored-reference rewind fix described in Sec.~\ref{sec:alignatt-llm},
the policy only reaches its upper quality bound at ${\sim}3.6$\,s of
CU-LongYAAL.

\subsection{End-to-End Cascade Quality and Latency}
\label{sec:exp-e2e}

Table~\ref{tab:e2e} reports the end-to-end cascade (Qwen3 forced ASR
$+$ Gemma AlignAtt MT, chunked synchronously) for the three IWSLT~2026
dev-set target languages. We report a low-latency regime at
$\Delta_{\mathrm{chunk}}{=}850$\,ms and a single quality-anchor row
per pair at $\Delta_{\mathrm{chunk}}{=}1500$\,ms.

\begin{table}[t]
\centering
\small
\setlength{\tabcolsep}{3.5pt}
\begin{tabular}{l l c c c r r}
\toprule
Pair & Regime & BLEU & chrF & COMET & \multicolumn{2}{c}{LongYAAL (s)} \\
\cmidrule(lr){6-7}
& & & & & CU & CA \\
\midrule
\multirow{2}{*}{en$\to$de} & low     & $28.76$ & $62.1$ & $0.875$ & $2.00$ & $1.63$ \\
                           & quality & $32.63$ & $64.2$ & $0.902$ & $3.53$ & $3.14$ \\
\midrule
\multirow{2}{*}{en$\to$zh} & low     & $36.01$ & $35.0$ & $0.743$ & $1.95$ & $1.77$ \\
                           & quality & $39.86$ & $37.8$ & $0.778$ & $3.27$ & $3.09$ \\
\midrule
\multirow{2}{*}{en$\to$it} & low     & $40.10$ & $68.0$ & $0.805$ & $1.98$ & $1.62$ \\
                           & quality & $44.46$ & $70.1$ & $0.841$ & $3.48$ & $3.10$ \\
\bottomrule
\end{tabular}
\caption{\textbf{End-to-end cascade on IWSLT~2026 dev-set}
(21 clips, OmniSTEval longform resegmentation; COMET~=~XCOMET-XL).
Low regime uses $\Delta_{\mathrm{chunk}}{=}850$\,ms and the AlignAtt
policy of Sec.~\ref{sec:qk-fast}; quality regime uses
$\Delta_{\mathrm{chunk}}{=}1500$\,ms.}
\label{tab:e2e}
\end{table}

At ${\sim}2$\,s CU-LongYAAL all three pairs sit in a narrow band,
and the quality regime adds ${\sim}1.5$\,s of CU for ${+}3$--${+}4$
BLEU and ${+}0.02$--${+}0.04$ XCOMET-XL. The COMET gap between
en--de/en--it ($\geq 0.81$) and en--zh ($\leq 0.78$) is consistent
with the well-known distributional difference of Chinese references:
word-segmentation choices and the lower chrF~(${\leq}38$) magnify the
gap even at similar BLEU. Improving the en--zh prompt or introducing
target-language domain priors is left for future work.

% ============================================================
% SECTION 6 — DISCUSSION
% ============================================================
\section{Discussion}
\label{sec:discussion}

\paragraph{Single-substrate vs specialist cascades.}
The comparison in Sec.~\ref{sec:exp-asr} quantifies the engineering
cost of removing a dedicated aligner: roughly $2\times$ WER and
${\sim}0.25$\,s of additional boundary lag on the diagnostic clip,
against a halving of model footprint and the removal of an entire
runtime. In submissions where the MT model is already the dominant
cost, reusing its weights for ASR alignment via AlignAtt becomes
attractive, especially as the streaming frontier of
Sec.~\ref{sec:exp-frontier} exposes operating points below one second
of CU-LongYAAL that the specialist cascade cannot reach without
additional policy machinery on top of the forced aligner.

\paragraph{Numerical parity as an enabling condition.}
The parity bound of Eq.~\eqref{eq:parity}, together with the
reconstruction-error bound of Eq.~\eqref{eq:robust}, is what makes the
AlignAtt substrate operational under compiled inference. Without
bit-identical acceptance decisions, every ablation against a reference
Transformers forward would conflate policy effects with capture noise;
with them, every experiment in Sec.~\ref{sec:experiments} is carried
out on the engine-native path used in deployment.

\paragraph{Limitations and future work.}
Our evaluation is single-GPU and single-language-family (en source, one
each of de/zh/it target); the policy has not been stress-tested on
code-switching or long overlapping speakers. The cascade loop is
synchronous: ASR and MT chunks are serialized, so the CU-LongYAAL
numbers in Table~\ref{tab:e2e} embed a full chunk round-trip rather
than a pipelined one. Finally, the en--zh quality gap suggests that the next
lever is on the MT side (larger base model, prompt engineering with
target-language priors), not on AlignAtt. AlignAtt itself, as a policy
substrate, is not the bottleneck on any pair tested.

\begin{thebibliography}{9}

\bibitem{alignatt23}
Sara Papi, Marco Turchi, and Matteo Negri.
\newblock AlignAtt: Using Attention-based Audio-Translation Alignments as a
Guide for Simultaneous Speech Translation.
\newblock In \emph{Proceedings of Interspeech 2023}, 2023.

\bibitem{tokenheads26}
Binbin Liu, Wenhan Han, Feng Chen, Yifan Zhang, Ping Guo, Haobin Lin, Bingni
Zhang, Taifeng Wang, and Yin Zheng.
\newblock Token Alignment Heads: Unveiling Attention's Role in LLM Multilingual
Translation.
\newblock \emph{ICLR 2026 Poster}, OpenReview, 2026.

\bibitem{streamatt24}
Sara Papi, Marco Gaido, Matteo Negri, and Luisa Bentivogli.
\newblock StreamAtt: Direct Streaming Speech-to-Text Translation with
Attention-based Audio History Selection.
\newblock In \emph{Proceedings of ACL 2024}, 2024.

\bibitem{simulstream25}
Marco Gaido, Sara Papi, Mauro Cettolo, Matteo Negri, and Luisa Bentivogli.
\newblock Simulstream: Open-Source Toolkit for Evaluation and Demonstration of
Streaming Speech-to-Text Translation Systems.
\newblock arXiv preprint arXiv:2512.17648, 2025.

\bibitem{longyaal25}
Peter Pol{\'a}k, Sara Papi, Luisa Bentivogli, and Ond{\v{r}}ej Bojar.
\newblock Better Late Than Never: Meta-Evaluation of Latency Metrics for
Simultaneous Speech-to-Text Translation.
\newblock arXiv preprint arXiv:2509.17349, 2025.

\bibitem{cuniiwslt25}
Dominik Mach{\'a}{\v{c}}ek and Peter Pol{\'a}k.
\newblock Simultaneous translation with offline speech and LLM models in CUNI
submission to IWSLT 2025.
\newblock In \emph{Proceedings of the 22nd International Conference on Spoken
Language Translation (IWSLT 2025)}, pages 389--398, Vienna, Austria (in-person
and online). Association for Computational Linguistics, 2025.

\bibitem{xcometxl23}
Nuno M. Guerreiro, Ricardo Rei, Daan van Stigt, Luisa Coheur, Pierre Colombo,
and Andr{\'e} F. T. Martins.
\newblock xCOMET: Transparent Machine Translation Evaluation through Fine-grained
Error Detection.
\newblock \emph{Transactions of the Association for Computational Linguistics},
12:979--995, 2024.

\bibitem{qwen3asr25}
Qwen Team.
\newblock Qwen3-ASR: A multilingual speech recognition model with
forced-alignment timestamp supervision.
\newblock Technical report, Alibaba Cloud, 2025.

\end{thebibliography}

\end{document}
